---
title: "Module 1 Deliverable"
author: "Joshua Patrick"
format: html
---

## Setup

```{r}
library(tidyverse)
dat <- read_csv("data/store_sales.csv")
```

## 1. Framing

Which regional manager should we promote based on revenue per square foot?

## 2. Data Description

The data consists of four months of performance across six retail stores. There are 24 observations and 6 variables. 

The variables are listed below.

| Column | Description | Variable Type |
|------------------------|------------------------|------------------------|
| `store_id` | Unique identifier for the store | Categorical |
| `region` | Region the store is located in (West, South, Midwest) | Categorical |
| `month` | Month of the observation (Jan–Apr) | Categorical |
| `sq_ft` | Store size, in square feet | Quantitative, continuous |
| `units_sold` | Number of units sold that month | Quantitative, discrete |
| `revenue` | Revenue that month, in dollars | Quantitative, discrete |

## 3. Table

```{r}
table(dat$region)
```


```{r}
dat |> 
  group_by(region) |> 
  summarize(avg_rev = mean(revenue))
```

```{r}
dat |> 
  group_by(region) |> 
  summarize(avg_sq_ft = mean(sq_ft))
```

## 4. Graphs

```{r}
dat |> 
  ggplot()+
  aes(x = region, y = revenue)+
  geom_point()
```

```{r}
dat |> 
  ggplot()+
  aes(x = region, y = sq_ft)+
  geom_point()
```

## 5. Descriptive Statistics

```{r}
median(dat$revenue)
sd(dat$revenue)
```

```{r}
dat |> 
  group_by(region) |> 
  summarize(md_rev = median(revenue))
```

## 6 Executive Summary
