---
title: "Module 1, Part 3: Practice Examples"
subtitle: "Tables, graphs, descriptive statistics, and the executive summary"
author: "QBA 5330"
format:
  html:
    theme: cosmo
    toc: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---



## Example 3A: New Dataset, Same Skills

```{r}
library(tidyverse)
ns <- read_csv("data/nutrition_survey.csv")
glimpse(ns)
```

**Your turn:** build each of the following.

**1. A frequency table for `region`.**
```{r}
# your code here

```

**2. A grouped summary table: mean `servings_per_day` by `region`.**
```{r}
# your code here

```

**3. A histogram of `household_income`.**
```{r}
# your code here

```

**4. Two descriptive statistics, correctly matched to variable type**: the mean *and* median of `household_income` (quantitative), and the proportion of respondents in one region of your choice (categorical).
```{r}
# your code here

```

## Example 3B: Executive Summary Sprint (5 minutes)

Using your own output from 3A, write one polished executive-summary paragraph (3–6 sentences), modeled on the strong-vs-weak example from earlier in this session: interpret your numbers, don't just restate them, and tie them to a plausible business decision.

*Write your executive summary here.*

**Pairs:** swap with a neighbor. Check their paragraph against two questions only: does it interpret rather than restate a number, and does it tie to a decision?

## Example 3C: Group Debug Sprint, Fix the Broken Report

The code and sentence below are meant to describe `data/training_program_anova.csv`, but they contain **three errors**. Find and fix all three.

```{r}
#| eval: false
library(tidyverse)
train <- read_csv("data/training_program_anova.csv")

ggplot(train, aes(x = row_number(), y = score)) +
  geom_col()

mean(train$program)
```

> "The average score was calculated."

**Your turn:** first, load the data so you can actually test your fixes.
```{r}
train <- read_csv("data/training_program_anova.csv")
glimpse(train)
```

**Error 1**, a graph problem: what's wrong with the plot above, and what should it be instead? Write the corrected code.
```{r}
# your corrected code here

```

**Error 2**, a statistic problem: what's wrong with `mean(train$program)`, and what should be computed instead? Write the corrected code.
```{r}
# your corrected code here

```

**Error 3**, a writing problem: rewrite the executive-summary sentence so it actually interprets a result rather than just stating that a calculation happened. Use real numbers from your fixes above.

*Write your corrected sentence here.*

## The Module 1 Deliverable

A short Quarto exploratory report, containing:

- **Framing** (Part 1)
- **Data description**, observational unit + at least 4 variables classified (Part 1)
- **Executive summary**
- At least one **table**
- At least one **graph**
- At least two **descriptive statistics**, correctly matched to variable type

You just practiced every one of these, across all three parts of today's session, on real data. `store_sales.csv`, or a dataset of your own from your workplace (instructor approval required), due before the next module.
