---
title: "Tables, Graphs, and Descriptive Statistics"
subtitle: "QBA 5330 — Module 1, Part 3 (September 4)"
author: "QBA 5330"
format:
  revealjs:
    theme: simple
    slide-number: true
    footer: "QBA 5330 | Module 1 | Part 3"
    transition: fade
    incremental: true
execute:
  echo: true
  warning: false
  message: false
---

## 

We know how to frame a decision, and where trustworthy data comes from.

**Now:** how do we actually look at it?

## Matching the Graph to the Variable

- **Categorical** variable &rarr; frequency table &rarr; **bar chart**
- **Quantitative** variable &rarr; binned counts &rarr; **histogram**

## Bar Chart: `region` (Categorical)

```{r}
#| echo: true
#| message: false
library(tidyverse)
sales <- read_csv("store_sales.csv")

sales |>
  count(region) |>
  ggplot(aes(x = region, y = n)) +
  geom_col() +
  labs(title = "Store-months by region", x = "Region", y = "Count")
```

## Histogram: `revenue` (Quantitative)

```{r}
#| echo: true
ggplot(sales, aes(x = revenue)) +
  geom_histogram(bins = 8) +
  labs(title = "Distribution of monthly revenue", x = "Revenue ($)", y = "Count")
```

## A Common Mix-Up

Building a bar chart of a **quantitative** variable's raw values (one bar per row) instead of a histogram.

::: {.notes}
Show this mistake deliberately, then show the corrected histogram, so the failure mode is memorable rather than abstract.
:::

## A Richer Table: Grouped Summary

```{r}
#| echo: true
sales |>
  group_by(region) |>
  summarize(mean_revenue = mean(revenue), n = n())
```

**Not just how many — how they *performed*.**

::: {.notes}
The rubric explicitly accepts a table of summary statistics broken out by group as an alternative to a plain frequency table — and it's usually more decision-relevant.
:::

## Center and Spread

- **Mean** — arithmetic average; sensitive to extreme values
- **Median** — middle value when sorted; resistant to extreme values
- **Standard deviation** — typical distance from the mean

```{r}
#| echo: true
sales |>
  summarize(
    mean_revenue = mean(revenue),
    median_revenue = median(revenue),
    sd_revenue = sd(revenue)
  )
```

## When the Mean Misleads

Add one hypothetical flagship store with unusually high revenue...

- **Mean** shifts noticeably
- **Median** barely moves

> Means are pulled toward extreme values; medians are not.

::: {.notes}
Preview, without going deep: this same center-and-spread idea returns in Module 2 as the foundation for confidence intervals — "spread" here is the intuitive ancestor of "standard error" later.
:::

## A Mistake Worth Making on Purpose

```{r}
#| echo: true
#| eval: false
mean(sales$region)
```

**"Average region" isn't a real quantity.**

::: {.notes}
Run this live and let it visibly fail or return NA with a warning. This is the rubric's own instruction — "do not report a mean for a categorical variable" — made concrete.
:::

## A Statistic That Actually Fits

```{r}
#| echo: true
sales |>
  count(region) |>
  mutate(proportion = n / sum(n))
```

*"West stores account for about a third of all store-months."*

**Never a mean, for a categorical variable. A proportion or a mode instead.**

## Executive Summary: Strong vs. Weak

Worth **20 of 100 points** — the single largest line item.

**Weak** (restates a number):

> "Average revenue was \$65,000."

**Strong** (interprets it, tied to a decision):

> "Midwest stores generated noticeably less revenue per month than South or West stores, even after accounting for smaller store size, worth investigating before allocating next quarter's inventory evenly across regions."

::: {.notes}
The weak version could describe almost any numeric column. The strong version connects a specific pattern in this data to a specific decision — the same framing skill from Part 1, now applied to results instead of a starting question.
:::

## Your Turn: One Sentence (2 minutes)

Using the grouped table from earlier, write **one sentence** that could open your own executive summary.

**Model it on the strong example — not the weak one.**

## Your Turn: Putting It All Together

Using `store_sales.csv`, individually:

1. **One table** — frequency table or grouped summary
2. **One graph** — a histogram of `units_sold`
3. **Two statistics, matched to type** — mean *and* median of `units_sold` (quantitative), *and* a proportion for one region (categorical)
4. **One sentence** — in the style of the strong executive summary

::: {.notes}
A compressed, ungraded dry run of the entire deliverable structure. The goal is to surface software friction and writing friction — slipping back into "weak" summary language under time pressure — while help is still in the room.
:::

## The Module 1 Deliverable

A short Quarto exploratory report, containing:

- **Framing** (Part 1)
- **Data description** — observational unit + at least 4 variables classified (Part 1)
- **Executive summary**
- At least one **table**
- At least one **graph**
- At least two **descriptive statistics**, correctly matched to variable type

## You Already Practiced All of It

- `store_sales.csv`, or a dataset of your own from your workplace (instructor approval required)
- Every skill above: practiced today, across all three parts

**Due before the next module.**

