---
title: "Module 1, Part 3: Practice Example Solutions"
subtitle: "Instructor answer key: tables, graphs, descriptive statistics, and the executive summary"
author: "QBA 5330"
format:
  html:
    theme: cosmo
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

::: {.callout-note}
This is an **instructor answer key** for the three Part 3 practice examples (3A, 3B, 3C), not a student handout. Every number, table, and plot below comes from actually running the code against the real files in `data/`.
:::

## Example 3A: New Dataset, Same Skills

```{r}
library(tidyverse)
ns <- read_csv("data/nutrition_survey.csv")
glimpse(ns)
```

**Table 1: frequency table for `region`**
```{r}
ns |> count(region)
```

**Table 2: a grouped summary (mean `servings_per_day` by `region`)**
```{r}
ns |> group_by(region) |> summarize(mean_servings = mean(servings_per_day), n = n())
```

**Graph: histogram of `household_income`**
```{r}
ggplot(ns, aes(x = household_income)) +
  geom_histogram(bins = 15, fill = "#2c7fb8", color = "white") +
  labs(title = "Distribution of household income", x = "Household income ($)", y = "Count")
```

**Quantitative statistics for `household_income`**
```{r}
ns |> summarize(mean_income = mean(household_income),
                 median_income = median(household_income),
                 sd_income = sd(household_income))
```

**Categorical statistic for `region`**
```{r}
ns |> count(region) |> mutate(proportion = n / sum(n))
```

## Example 3B: Executive Summary Sprint

Using the real output from 3A:

> Average household income across the 200 surveyed households was about \$54,600 (median \$53,700), with a typical household's income varying from that average by roughly \$11,900. Average daily fruit-and-vegetable servings were fairly similar across all four regions, ranging narrowly from about 3.0 servings in the South to 3.4 in the North, a gap of less than half a serving. Because no single region stands out as driving meaningfully healthier eating habits, a marketing campaign built around regional targeting would have little real signal to work with here; a national approach, or one built around income rather than region, looks more promising.

**What this models for students:** every sentence uses a number that was actually computed above, and every sentence does something *with* that number (compares it, questions whether it's decision-relevant, states an implication) rather than simply reporting it. Note also that this is an honest "we didn't find much of a regional story" summary, not a forced dramatic finding, and that's still a strong, usable executive summary.

## Example 3C: Group Debug Sprint, Fix the Broken Report

### The broken starter (hand this out or project it)

```{r}
#| eval: false
library(tidyverse)
train <- read_csv("data/training_program_anova.csv")

ggplot(train, aes(x = row_number(), y = score)) +
  geom_col()

mean(train$program)
```

> "The average score was calculated."

### The three errors, and the fix

```{r}
train <- read_csv("data/training_program_anova.csv")
glimpse(train)
```

**Error 1: a bar chart of a quantitative variable's raw values, instead of a histogram.**
```{r}
ggplot(train, aes(x = score)) +
  geom_histogram(bins = 12, fill = "#2c7fb8", color = "white") +
  labs(title = "Distribution of training assessment scores", x = "Score", y = "Count")
```

**Error 2: `mean()` computed on a categorical variable.** `program` (A, B, or C) has no numeric meaning to average; the fix is a frequency table or a proportion, not a mean.
```{r}
train |> count(program) |> mutate(proportion = n / sum(n))
```

**Error 3: a weak executive-summary sentence that restates a calculation instead of interpreting it.**
```{r}
train |> summarize(mean_score = mean(score), median_score = median(score), sd_score = sd(score))
```

> **Weak (what was handed out):** "The average score was calculated."
>
> **Fixed:** "Average assessment scores across all 150 participants were about 76 points (median 76.3, typical variation of about 5.2 points), with Program C the most commonly completed program (59 of 150 participants, about 39%). Because scores are fairly tightly clustered around the average regardless of program, there's no obvious sign yet that one program produces meaningfully different results, worth confirming with a formal comparison (an ANOVA, covered later in this course) before recommending that the company favor one program over another."

**What to check when grading a group's fix:** did they change the *graph type* (not just relabel the broken one), did they replace the mean with a statistic that actually applies to a categorical variable, and does the rewritten sentence interpret the numbers rather than just describe that a calculation happened?
