---
title: "Quantifying Uncertainty"
subtitle: "QBA 5330 — Module 2, Part 1 (October 16)"
author: "QBA 5330"
format:
  revealjs:
    theme: simple
    slide-number: true
    footer: "QBA 5330 | Module 2 | Part 1"
    transition: fade
    incremental: true
execute:
  echo: true
  warning: false
  message: false
---

## Where We're Headed Today

- **Part 1:** Quantifying uncertainty, from probability to sampling distributions
- **Part 2:** Confidence intervals for proportions and means
- **Part 3:** Introduction to hypothesis testing, and launching the Module 2 deliverable

::: {.notes}
Module 1 handled framing, obtaining, and summarizing data. The five weeks since 9/4 built the entire toolkit for quantifying uncertainty. Today we use it, on one running business decision.
:::

## Today's Business Decision

> A retailer has piloted a new online storefront in a sample of markets. Should it launch nationally?

We'll carry this same decision through all three parts.

## A Quick Example

Leadership estimates:

- National launch cost: **\$2,000,000**
- Expected incremental profit: **\$2,600,000**
- Standard deviation of that profit estimate: **\$1,800,000**

- **Is this a "safe bet"?**

::: {.notes}
Expected value alone says yes (margin of $600,000), but the standard deviation is large relative to that margin, so the chance of a loss is far from zero. This is why we don't stop at a point estimate.
:::

## Recap: Expected Value and Risk

- Expected value is a **long-run average**, not a guarantee for any one outcome
- A positive expected value can still carry real risk
- We need both the **center** (expected value) and the **spread** (standard deviation) to reason about risk

## Recap: Sampling Distributions

- A sample mean or sample proportion is a **statistic**
- Every statistic has a **sampling distribution**: how much it would vary across repeated samples
- The **standard error** is the standard deviation of that sampling distribution

## The Two Standard Error Formulas

$$
\text{SE}_{\bar{x}} = \frac{s}{\sqrt{n}} \qquad\qquad \text{SE}_{\hat{p}} = \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}
$$

A bigger sample doesn't eliminate uncertainty. It **shrinks the standard error**.

## The Pilot Data

- **Conversion:** $n=200$ visitors, 30 converted, $\hat{p} = 0.15$
- **Order value:** $n=25$ orders, $\bar{x} = \$71.50$, $s = \$42$

*(You've seen these numbers before, in your Week 4–6 reading.)*

## Your Turn (20 minutes)

In your group, compute both standard errors:

```{r}
#| echo: true
p_hat <- 0.15; n_conv <- 200

```



```{r}
#| echo: true
s <- 42; n_order <- 25

```

**Then discuss:** what does each number mean for the launch decision?

## Before the Break

We now have:

- A point estimate ($\hat{p}=0.15$, $\bar{x}=\$71.50$)
- A sense of how much each could bounce around (the standard errors)

- **What we don't have yet:** a defensible *range* to bring to leadership.

- That's next.
