---
title: "Hypothesis Testing and the Module 2 Deliverable"
subtitle: "Module 2, Part 3 (October 16)"
author: "QBA 5330"
format:
  revealjs:
    theme: simple
    slide-number: true
    footer: "QBA 5330 | Module 2 | Part 3"
    transition: fade
    incremental: true
execute:
  echo: true
  warning: false
  message: false
---

## A Different Question

- Part 2 asked: *is the true national rate at least leadership's minimum bar?*

- New question: *has the new storefront's performance actually changed from the old storefront's history?*

- That's a **hypothesis test** question.

## Testing the Conversion Rate

$H_0: p = 0.12$ (historical rate) vs. $H_a: p \neq 0.12$

```{r}
#| echo: true
p_hat <- 0.15; p0 <- 0.12; n <- 200

z_stat <- (p_hat - p0) / sqrt(p0 * (1 - p0) / n)
p_value <- 2 * (1 - pnorm(abs(z_stat)))
c(z = z_stat, p_value = p_value)
```

At $\alpha = 0.05$: **fail to reject** $H_0$.

::: {.notes}
Reconnect to Part 2: the 95% CI (0.101, 0.199) contains 0.12, so the test and interval agree.
:::

## Testing the Order Value

$H_0: \mu = 65$ (historical average) vs. $H_a: \mu \neq 65$

```{r}
#| echo: true
x_bar <- 71.50; mu0 <- 65; s <- 42; n <- 25

t_stat <- (x_bar - mu0) / (s / sqrt(n))
p_value <- 2 * pt(-abs(t_stat), df = n - 1)
c(t = t_stat, p_value = p_value)
```

At $\alpha = 0.05$: **fail to reject** $H_0$.

## Three Lenses on One Pilot {.smaller}

| Question | Tool | This pilot |
|---|---|---|
| Best single guess? | Point estimate | $\hat{p}=0.15$; $\bar{x}=\$71.50$ |
| Plausible range? | Confidence interval | $(0.101,0.199)$; $(\$54.16,\$88.84)$ |
| Clears leadership's bar? | CI vs. threshold | Conversion: barely. Order value: not clearly. |
| Changed from history? | Hypothesis test | Fail to reject, both |

## The Honest Takeaway

No clean "yes, launch" or "no, don't."

Good analytics **narrows the range of reasonable disagreement**. It doesn't always resolve it.

## The Module 2 Deliverable

- A short **risk and uncertainty exercise**, using the same three-lens approach:

1. Expected value / probability
2. Confidence interval
3. Hypothesis test

- Applied to a scenario involving **demand, quality, conversion, or customer behavior**, calculated in Excel and reproduced or extended in R.

## Looking Ahead to Saturday

- Today: **quantifying** uncertainty.

- Tomorrow (10/17): **comparing groups** and **evaluating causal claims**, using the Google Project Oxygen case.


