---
title: "Module 2, Part 3: Hypothesis Tests and Your Deliverable"
subtitle: "Guided notebook: review the reading, then fill in the blanks together"
author: "Your name here"
format:
  html:
    theme: cosmo
    toc: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
---

::: {.callout-important title="Reminder: how this notebook works"}
- Replace each `___` in code with the missing R code, then run the chunk. Type your interpretation where you see `_____` in a sentence.
- Run chunks **in order from the top**. Later chunks use objects created by earlier ones.
- Render any time. Unfilled blanks show red error text but do not stop the render.
- **Exception:** an error in R code written *inside a sentence* (inline code) does stop the render. Fix or remove it, then render again.
:::

## Setup and Recap from Parts 1 and 2

Run this chunk. It loads the data and recreates the numbers and intervals from the first two parts.

```{r}
library(tidyverse)

base_url <- "https://raw.githubusercontent.com/joshdp34/QBA/master/data/"
pilot_conversions <- read_csv(paste0(base_url, "retailer_pilot_conversions.csv"))
pilot_orders      <- read_csv(paste0(base_url, "retailer_pilot_orders.csv"))

# Conversion rate
n_conv <- nrow(pilot_conversions)
x_conv <- sum(pilot_conversions$converted)
p_hat  <- mean(pilot_conversions$converted)
se_p   <- sqrt(p_hat * (1 - p_hat) / n_conv)

# Order value
n_ord <- nrow(pilot_orders)
x_bar <- mean(pilot_orders$order_value)
s     <- sd(pilot_orders$order_value)
se_x  <- s / sqrt(n_ord)

# 95% intervals from Part 2
ci_p <- c(lower = p_hat - qnorm(0.975) * se_p, upper = p_hat + qnorm(0.975) * se_p)
ci_x <- c(lower = x_bar - qt(0.975, n_ord - 1) * se_x, upper = x_bar + qt(0.975, n_ord - 1) * se_x)

round(ci_p, 3)
round(ci_x, 2)
```

## 1. A Different Question (review)

**Key ideas from the reading** (Weeks 5 and 6 readings):

- A **confidence interval** answers: *what range of values is the truth plausibly in?*
- A **hypothesis test** answers: *has the truth changed from a known benchmark?*

Every test follows the same five steps:

1. State $H_0$ (nothing has changed) and $H_a$ (something has).
2. Compute a **test statistic**: how many standard errors the pilot is from the benchmark.
3. Find the **p-value**: how likely a result at least this extreme would be if $H_0$ were true.
4. Compare the p-value to the significance level $\alpha$ (usually 0.05). If p-value $< \alpha$, **reject** $H_0$. Otherwise, **fail to reject** $H_0$.
5. Say it in plain business language.

::: {.callout-note title="'Fail to reject' is not 'accept'"}
Failing to reject $H_0$ means the pilot did not provide *statistically detectable* evidence of a change. It does not prove that nothing changed. A larger pilot might detect a smaller difference.
:::

## 2. Test 1: Has the Conversion Rate Changed?

The **old** storefront historically converted **12%** of visitors. The pilot converted 15%.

$$
H_0: p = 0.12 \qquad H_a: p \neq 0.12
$$

Fill in the blanks to compute the test statistic and p-value by hand.

```{r}
p0     <- ___                                  # the historical conversion rate
se_0   <- sqrt(p0 * (1 - p0) / n_conv)              # standard error UNDER H0
z_stat <- (p_hat - ___) / ___
p_value <- 2 * (1 - pnorm(abs(___)))         # two-sided p-value
c(z = z_stat, p_value = p_value)
```

::: {.callout-tip collapse="true" title="Why does the standard error use p0 instead of p_hat?"}
A hypothesis test asks "if $H_0$ were true, how surprising is the pilot?" So the standard error is computed *assuming $H_0$ is true*, using the benchmark $p_0$. A confidence interval has no benchmark, so it uses $\hat{p}$ instead.
:::

Check your work with R's built-in test. The `X-squared` value equals $z^2$, and the p-value should match yours:

```{r}
prop.test(x = x_conv, n = n_conv, p = 0.12, correct = FALSE)
```

Now make the decision:

```{r}
alpha <- 0.05
p_value < alpha
```

**Interpretation.** Fill in the blanks (the first two use inline code):

> The test statistic is $z$ = `r "____"` and the p-value is `r "____"`. Because the p-value is _____ than $\alpha = 0.05$, we _____ $H_0$. The pilot does not provide statistically detectable evidence that the new storefront's conversion rate differs from the old storefront's 12%.

**Connect it to Part 2:** the 95% confidence interval for the conversion rate was about (0.101, 0.199). The value 0.12 is _____ that interval, which _____ with the test's conclusion. A confidence interval and a two-sided test at matching levels always tell the same story.

## 3. Test 2: Has the Average Order Value Changed?

The old storefront's historical average order value was **\$65**. The pilot averaged \$71.50.

$$
H_0: \mu = 65 \qquad H_a: \mu \neq 65
$$

For a mean with raw data, R's `t.test()` does the whole test. Fill in the benchmark:

```{r}
order_test <- t.test(pilot_orders$order_value, mu = ___)
order_test
```

You can pull individual results out of the saved test. Fill in the blank:

```{r}
order_test$statistic     # the t statistic
order_test$___   # the p-value
```

Now confirm the t statistic **by hand**. It is the same recipe: (estimate minus benchmark) divided by the standard error.

```{r}
mu0    <- 65
t_stat <- (x_bar - ___) / ___
p_val  <- 2 * (1 - pt(abs(t_stat), df = ___))
c(t = t_stat, p_value = p_val)
```

**Interpretation.** Fill in the blanks:

> The test statistic is $t$ = `r "____"` with a p-value of `r "____"`. Because the p-value is _____ than $\alpha = 0.05$, we _____ $H_0$. The pilot's average order value (\$71.50) is not statistically distinguishable from the old storefront's \$65. The value 65 is _____ the 95% interval (\$54.16, \$88.84) from Part 2.

### What if we change $\alpha$?

The significance level is a choice, and it controls how often we raise a false alarm (a Type I error). Check the conversion test at two other levels. Fill in the blanks.

```{r}
p_value < ___
p_value < ___
```

At $\alpha = 0.20$, we would _____ $H_0$ for the conversion test. But that means accepting a _____% chance of a false alarm when nothing has really changed. At $\alpha = 0.01$ we would _____.

## 4. Three Lenses on One Pilot

Complete the last column. Use short, plain-language phrases.

| Question | Tool | Answer for this pilot |
|---|---|---|
| What is our best single guess? | Point estimate | Conversion _____; average order value _____ |
| What range is the truth plausibly in? | Confidence interval | Conversion _____; order value _____ |
| Does the truth clear leadership's minimum bar? | CI vs. threshold | Conversion _____; order value _____ |
| Has performance changed from the old storefront? | Hypothesis test | _____ |

::: {.callout-tip title="The honest takeaway"}
None of these tools hands leadership a clean "yes, launch" or "no, don't." The pilot has not been shown to perform *differently* from the old storefront, and order value in particular cannot be confidently said to clear the \$60 bar. That is a realistic outcome, not a disappointing one: good analytics **narrows the range of reasonable disagreement**. A larger pilot might settle the question either way.
:::

## 5. You Are Ready for the Deliverable

Across all three parts you have now done every calculation the Module 2 deliverable asks for. Here is the map:

| Deliverable section | Where you practiced it |
|---|---|
| 1. Framing | Module 1 skills, plus the business decision at the top of Part 1 |
| 2. Probability and expected value | Part 1, Section 1 (`pnorm()` for the probability of a loss) |
| 3. Confidence interval | Part 2 (change `level` to the level the assignment asks for) |
| 4. Hypothesis test | Part 3, Sections 2 and 3 (use the $\alpha$ that matches your interval's level) |
| 5. Recommendation | Part 3, Section 4 (the honest takeaway, in your own words) |

Open **`mod2_deliverable_template.qmd`**. Its structure matches the assignment, with a reminder at each step of which part of these notebooks to look back at.

## Recap

| Task | Tool in R |
|---|---|
| Test statistic for a proportion | `(p_hat - p0) / sqrt(p0 * (1 - p0) / n)` |
| Two-sided p-value from a z statistic | `2 * (1 - pnorm(abs(z)))` |
| Whole test for a proportion | `prop.test(x, n, p = p0, correct = FALSE)` |
| Whole test for a mean | `t.test(x, mu = mu0)` |
| Pull the p-value out of a saved test | `result$p.value` |
| Decision | reject $H_0$ if `p_value < alpha` |
