---
title: "Confidence Intervals for Proportions and Means"
subtitle: "Module 2, Part 2 (October 16)"
author: "QBA 5330"
format:
  revealjs:
    theme: simple
    slide-number: true
    footer: "QBA 5330 | Module 2 | Part 2"
    transition: fade
    incremental: true
execute:
  echo: true
  warning: false
  message: false
---

## Where We Left Off

We have a point estimate and a standard error for the pilot launch data.

**Missing:** a defensible range to bring to leadership.

$$
\text{point estimate} \;\pm\; \text{critical value} \times \text{standard error}
$$

## Confidence Interval for the Conversion Rate

$\hat{p} = 0.15$, $n = 200$, 95% confidence

```{r}
#| echo: true
p_hat <- 0.15
n <- 200
se <- sqrt(p_hat * (1 - p_hat) / n)
z <- qnorm(0.975)

c(lower = p_hat - z * se, upper = p_hat + z * se)
```

*"We're 95% confident the true national conversion rate is between about 10.1% and 19.9%."*

## Confidence Interval for the Order Value

$\bar{x} = \$71.50$, $s = \$42$, $n = 25$, 95% confidence

```{r}
#| echo: true
x_bar <- 71.50
s <- 42
n <- 25
se <- s / sqrt(n)
t <- qt(0.975, df = n - 1)

c(lower = x_bar - t * se, upper = x_bar + t * se)
```

::: {.notes}
Flag explicitly: this uses a t critical value, not z, since sigma is unknown and estimated by s.
:::

## Your Turn: Does the Pilot Clear the Bar? (15 minutes)

Leadership says the launch is worth pursuing only if:

- True national conversion rate is **at least 10%**
- True national average order value is **at least \$60**

- **Check both intervals against these thresholds. Report out.**

## What Your Group Should Find

- Conversion: $(0.101, 0.199)$ — barely clears 10%
- Order value: $(54.16, 88.84)$ — **includes** values below \$60

::: {.notes}
This is a deliberately mixed signal. One metric barely clears the bar; the other's point estimate looks fine but its interval doesn't rule out falling short. A point-estimate-only report would have hidden this.
:::

## Before the Break

- Confidence intervals give us a **range** of plausible values.

- But leadership asked a **yes/no question**: does the average order value exceed \$60?

- Formalizing that yes/no question is exactly what a **hypothesis test** does.
