---
title: "Module 2, Part 2: Confidence Intervals for Proportions and Means"
subtitle: "Guided notebook: review the reading, then fill in the blanks together"
author: "Your name here"
format:
  html:
    theme: cosmo
    toc: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
---

::: {.callout-important title="Reminder: how this notebook works"}
- Replace each `___` in code with the missing R code, then run the chunk. Type your interpretation where you see `_____` in a sentence.
- Run chunks **in order from the top**. Later chunks use objects created by earlier ones.
- Render any time. Unfilled blanks show red error text but do not stop the render.
- **Exception:** an error in R code written *inside a sentence* (inline code) does stop the render. Fix or remove it, then render again.
:::

## Setup and Recap from Part 1

Run this chunk. It loads the pilot data and recreates every number we computed in Part 1, so you don't have to.

```{r}
library(tidyverse)

base_url <- "https://raw.githubusercontent.com/joshdp34/QBA/master/data/"
pilot_conversions <- read_csv(paste0(base_url, "retailer_pilot_conversions.csv"))
pilot_orders      <- read_csv(paste0(base_url, "retailer_pilot_orders.csv"))

# Conversion rate: point estimate and standard error
n_conv <- nrow(pilot_conversions)
p_hat  <- mean(pilot_conversions$converted)
se_p   <- sqrt(p_hat * (1 - p_hat) / n_conv)

# Order value: point estimate and standard error
n_ord <- nrow(pilot_orders)
x_bar <- mean(pilot_orders$order_value)
s     <- sd(pilot_orders$order_value)
se_x  <- s / sqrt(n_ord)

tibble(metric = c("Conversion rate", "Average order value"),
       estimate = c(p_hat, x_bar),
       standard_error = c(se_p, se_x))
```

We left Part 1 with a point estimate and a standard error for each metric, but no *range* leadership can rely on. That is what a confidence interval gives us.

## 1. The Confidence Interval Recipe (review)

**Key ideas from the reading** (Week 5 reading):

$$
\text{estimate} \;\pm\; \text{critical value} \times \text{standard error}
$$

- For a **proportion**, the critical value is a **z** value, found with `qnorm()`.
- For a **mean**, we use a **t** value with $n-1$ degrees of freedom, found with `qt()`, because the true standard deviation is unknown and estimated by $s$.
- The **confidence level** (95%, 90%, 99%) controls the critical value. More confidence means a larger critical value and a wider interval.

::: {.callout-note title="What '95% confident' does and does not mean"}
The method behind a 95% interval captures the true value in about 95% of repeated pilots. Once we compute **one** specific interval, the true value is either inside it or not. So we say "we are 95% confident the true rate is between..." and we do **not** say "there is a 95% probability the true rate is in this interval."
:::

## 2. Confidence Interval for the Conversion Rate

We store the confidence level in an object called `level`. That way, changing the level later means changing **one number**. The formula `1 - (1 - level) / 2` turns a level like 0.95 into the quantile 0.975 that `qnorm()` needs.

Fill in the four blanks to build the 95% interval.

```{r}
level  <- ___                                   # confidence level
z_star <- qnorm(1 - (1 - ___) / 2)             # critical value for the chosen level
margin <- z_star * ___                          # margin of error
ci_p   <- c(lower = p_hat - ___, upper = p_hat + ___)
ci_p
```

::: {.callout-tip collapse="true" title="Hint"}
The margin of error is the critical value times the standard error. The interval is the estimate minus the margin, and the estimate plus the margin.
:::

R also has a built-in function that computes a proportion interval. Run it and compare:

```{r}
prop.test(x = sum(pilot_conversions$converted), n = n_conv, correct = FALSE)$conf.int
```

The numbers are close but not identical, because `prop.test()` uses a slightly different (Wilson) method. For this course, **use the by-hand interval above** as your answer.

**Interpretation, using inline code.** Replace each placeholder with R code that pulls the endpoint from `ci_p` (multiply by 100 and round to 1 decimal):

> We are 95% confident that the true national conversion rate is between `r "____"`% and `r "____"`%.

### What if we want more confidence?

Rebuild the interval at a **99%** confidence level. Only the first line changes.

```{r}
level_99  <- ___
z_star_99 <- qnorm(1 - (1 - level_99) / 2)
margin_99 <- z_star_99 * se_p
ci_p_99   <- c(lower = p_hat - margin_99, upper = p_hat + margin_99)
ci_p_99
```

The 99% interval is _____ than the 95% interval. To be more confident that we have captured the true rate, we have to _____.

## 3. Confidence Interval for the Average Order Value

Because we are estimating a **mean** with $s$, we use the **t** distribution with $n - 1$ degrees of freedom. Fill in the four blanks.

```{r}
level    <- 0.95
df       <- ___                                    # degrees of freedom
t_star   <- qt(1 - (1 - level) / 2, df = ___)             # t critical value
margin_x <- ___ * se_x                                # margin of error
ci_x     <- c(lower = x_bar - ___, upper = x_bar + margin_x)
ci_x
```

R's `t.test()` computes the same interval directly. This time the match is exact:

```{r}
t.test(pilot_orders$order_value)$conf.int
```

**Interpretation, using inline code.** Dollar signs are written `\$` in Quarto sentences.

> We are 95% confident that the true national average order value is between \$`r "____"` and \$`r "____"`.

::: {.callout-tip collapse="true" title="Hint"}
Same pattern as before: `format(round(ci_x["lower"], 2), nsmall = 2)` keeps two decimals on a dollar amount.
:::

## 4. Does the Pilot Clear Leadership's Bar?

Leadership will pursue the national launch only if they can be reasonably confident that:

- the true national **conversion rate is at least 10%**, and
- the true national **average order value is at least \$60**.

The key question is whether the **lower end** of each interval clears the bar. Fill in the two thresholds.

```{r}
ci_p["lower"] > ___     # conversion bar
ci_x["lower"] > ___       # order value bar
```

Run this chunk (no changes needed) to *see* both intervals against leadership's bar (the red dashed line):

```{r}
tibble(metric   = c("Conversion rate", "Average order value ($)"),
       estimate = c(p_hat, x_bar),
       lower    = c(ci_p["lower"], ci_x["lower"]),
       upper    = c(ci_p["upper"], ci_x["upper"]),
       bar      = c(0.10, 60)) |>
  ggplot(aes(x = metric, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.15) +
  geom_hline(aes(yintercept = bar), linetype = "dashed", color = "red") +
  facet_wrap(~ metric, scales = "free") +
  labs(title = "95% confidence intervals vs. leadership's minimum bar",
       x = NULL, y = NULL) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

**Interpretation.** Circle or type the right words in each blank:

- **Conversion rate:** the lower bound of the interval is `r "____"`%, which _____ the 10% bar, but only just. Leadership should ask how comfortable they are with "just barely."
- **Order value:** the lower bound is \$`r "____"`, which _____ the \$60 bar. The pilot data _____ a true average order value below \$60, even though the point estimate of \$71.50 looks comfortably above it.

::: {.callout-note title="Why the interval matters"}
A report that only said "average order value was \$71.50, comfortably above \$60" would have hidden real cause for caution. Reporting the interval is what exposes it.
:::

## Recap

| Task | Tool in R |
|---|---|
| z critical value for a chosen `level` | `qnorm(1 - (1 - level) / 2)` |
| t critical value for a chosen `level` | `qt(1 - (1 - level) / 2, df = n - 1)` |
| Margin of error | critical value `*` standard error |
| Interval | `estimate - margin`, `estimate + margin` |
| Built-in check for a mean | `t.test(x)$conf.int` |
| Change the confidence level | change **one number**: `level` |

Intervals give a *range* of plausible values. But leadership also asked yes/no questions, such as "has the new storefront actually changed performance from the old one?" That is what a **hypothesis test** answers. See you in Part 3.
