---
title: "Module 3, Part 3: Project Oxygen, Causation, and Your Midterm"
subtitle: "Guided notebook: review the reading, then fill in the blanks together"
author: "Your name here"
format:
  html:
    theme: cosmo
    toc: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: true
---

::: {.callout-important title="Reminder: how this notebook works"}
- Replace each `___` in code with the missing R code, then run the chunk. Type your answer where you see `_____` in a sentence.
- Run chunks **in order from the top**. Later chunks use objects created by earlier ones.
- Render any time. Unfilled blanks show red error text but do not stop the render.
- **Exception:** an error in R code written *inside a sentence* (inline code) does stop the render. Fix or remove it, then render again.
- In Section 4 you write your **team's** answers to the worksheet. Sentences are fine; there is no wrong place to type.
:::

## Setup

```{r}
library(tidyverse)

base_url <- "https://raw.githubusercontent.com/joshdp34/QBA/master/data/"

manager_teams <- read_csv(paste0(base_url, "manager_teams.csv")) |>
  mutate(manager_group = factor(manager_group,
                                levels = c("Top quartile", "Bottom quartile")))

glimpse(manager_teams)
```

::: {.callout-warning title="About this data"}
Google's Project Oxygen did **not** release row-level data. The dataset above is **simulated** to have the same *structure* as the case: managers in the top or bottom quartile of manager scores, and an engagement score (0 to 100) for each manager's team. It is for practicing the reasoning, not a recreation of Google's results.
:::

| Column | Meaning |
|---|---|
| `manager_id` | Identifier for the manager |
| `manager_group` | Top quartile or Bottom quartile on the manager score |
| `team_experience` | Whether the manager's team is mostly Experienced or Newer employees |
| `engagement` | The team's average engagement score (0 to 100) |

## 1. The Case and the Comparison (review)

**Key ideas from the reading** (Week 7 reading): Google's people analytics team asked whether **managers matter**. They compared teams led by top-quartile and bottom-quartile managers on satisfaction, retention, and performance. That is a two-group comparison, which is exactly the toolkit from Parts 1 and 2.

First, summarize the two groups. Fill in the blank.

```{r}
manager_teams |>
  group_by(manager_group) |>
  summarise(n = n(),
            mean_engagement = ___(engagement),
            sd_engagement   = sd(engagement))
```

Which tool applies? The groups are two separate sets of managers, so it is a _____ test. Run it. Because we set `Top quartile` as the first level, the difference is "Top minus Bottom".

```{r}
oxy_test <- t.test(___ ~ manager_group, data = manager_teams)
oxy_test
```

**Interpretation, using inline code.** `oxy_test$estimate` holds the two group means (Top first) and `oxy_test$conf.int` the interval for the difference:

> Teams under top-quartile managers averaged `r "____"` and teams under bottom-quartile managers averaged `r "____"`. The 95% confidence interval for the difference runs from `r "____"` to `r "____"` points.

Fill in the blanks:

The p-value is far _____ 0.05, so we _____ $H_0$: the gap is statistically significant. The gap is about 6 points on a 0 to 100 scale, and the spread within each group (the standard deviation) is about 8 points, so the gap is roughly _____ standard deviations. That is a _____ difference, not a trivial one.

But the reading's warning applies: a significant difference between groups does **not** by itself tell us that managers *cause* the difference.

## 2. The Causation Checklist (review)

**Key ideas from the reading** (Week 7 reading). Before claiming a causal effect, ask four questions:

1. Was there **random assignment**?
2. Is there a plausible **confounder**: something that affects both who ends up in a group *and* the outcome?
3. Could **bias** explain the pattern instead of a real effect?
4. Is the effect **statistically significant, and practically significant**?

In Project Oxygen, nobody was randomly assigned to be a top or bottom manager. The comparison is _____, so confounding is the main worry.

## 3. A Confounder You Can See in the Data

Suppose better-rated managers tend to be given **more experienced teams**. Experienced teams might be more engaged no matter who manages them. That would make top managers look better than they are.

Check it. What fraction of each group's teams are Experienced? Fill in the blank.

```{r}
manager_teams |>
  group_by(manager_group) |>
  summarise(pct_experienced = mean(team_experience == "___"))
```

About _____ of top-quartile managers lead experienced teams, versus about _____ of bottom-quartile managers. The two groups are **not** comparable to begin with.

Now compare the managers **within** each type of team, so that experience is held roughly constant. Run the first chunk as is, then fill in the blanks in the second.

```{r}
manager_teams |>
  group_by(team_experience, manager_group) |>
  summarise(n = n(), mean_engagement = mean(engagement), .groups = "drop")
```

```{r}
exp_test   <- t.test(engagement ~ manager_group,
                     data = filter(manager_teams, team_experience == "Experienced"))
newer_test <- t.test(engagement ~ manager_group,
                     data = filter(manager_teams, team_experience == "___"))

gap <- function(test) unname(test$estimate[1] - test$estimate[2])   # Top minus Bottom

round(c(overall     = gap(oxy_test),
        experienced = gap(___),
        newer       = gap(newer_test)), 2)
```

```{r}
round(c(experienced = exp_test$p.value, newer = newer_test$p.value), 4)
```

**Interpretation.** Fill in the blanks:

The overall gap was about _____ points, but *within* each type of team it shrinks to about _____ points. Team experience therefore accounts for roughly _____ of the raw gap. A gap of that size remains in both groups, and both p-values are below 0.05, so there is still _____ that top-quartile managers' teams are more engaged.

::: {.callout-tip title="The honest conclusion"}
We could only adjust for the one confounder we happened to measure. **Unmeasured** confounders (the manager's own tenure, the department, whether the team was already high-performing) could shrink the gap further. The remaining gap is *consistent with* managers mattering, but it is still an observational comparison, so it does not *prove* it. Only random assignment of the "treatment" can do that.
:::

## 4. Team Worksheet: Evaluate the Evidence

Work with your team. Write your team's answer to each item. One or two sentences is plenty. (The group-level numbers above are from simulated data, but your reasoning here is about the real case.)

**1. Random assignment.** Was there random assignment in Project Oxygen? What would a randomized version look like?

> Our answer: _____

**2. Confounders.** Name at least **one confounder not already in the reading** (the reading names team quality/resources and reverse causation).

> Our answer: _____

**3. Bias.** Top and bottom managers were defined by upward-feedback surveys plus performance reviews. What kind of bias could that introduce?

> Our answer: _____

**4. Statistical vs. practical significance.** The differences are described as reliable but sometimes small. What more would you want to know before a company-wide training rollout?

> Our answer: _____

**5. A stronger design.** Propose a *specific* design that would strengthen the causal claim. Say who is randomly assigned to what.

> Our answer: _____

**6. Team verdict.** Does your team think the evidence supports "managers matter"? What is the strongest reason for your position?

> Our answer: _____

## 5. Class Synthesis

After the report-outs, we will fill in the class's consolidated position:

- **Strongest confounders, named by several teams:** _____
- **Most concrete stronger design:** _____
- **Class verdict:** _____

## 6. Module 3 in One Table, and Your Midterm

Complete the "R function" column from memory (look back at Parts 1 and 2 if you need to).

| Question | Tool | R function | What to report |
|---|---|---|---|
| One group vs. a known benchmark | One-sample t-test | _____ | $t$, p-value, interval |
| Two separate groups | Welch two-sample t-test | _____ | $t$, p-value, interval for the difference |
| Same units measured twice | Paired t-test | _____ | $t$, p-value, interval for the mean difference |
| Three or more groups | One-way ANOVA | _____ | $F$, both df, p-value |
| Which groups differ | Tukey's HSD | _____ | adjusted p-values and intervals |

You now have every tool the **midterm** needs. Here is where each piece of the midterm was practiced:

| Midterm section | Practiced in |
|---|---|
| Setup: reading data from GitHub, renaming columns, dates | Part 1, Setup |
| Observational unit; looking at the data | Part 1, "Business Setting" |
| 3A: one-sample test | Part 1, Section 2 |
| 3B: two-sample comparison, Welch vs. pooled, interval for the difference | Part 1, Section 3 |
| 3C: paired comparison (build the pairs, then test) | Part 2, Section 2 |
| 3D: ANOVA and Tukey's HSD | Part 2, Sections 3 and 4 |
| 3E: evaluating causality | Part 3, Sections 2 to 4 |
| Executive recommendation | Modules 1 and 2 (honest about uncertainty, plain language) |

Open **`mod3_midterm_template.qmd`**. Its structure matches the midterm, with a reminder at each step of which part of these notebooks to look back at. The midterm uses a **different** dataset (the Sample Superstore data), but the setup code and every test work the same way. Your job is to swap in the right column names.

::: {.callout-tip title="Three habits that will earn points on the midterm"}
1. **Say which tool you chose and why** (one sentence). Naming the design is half the answer.
2. **Report the size of a difference, not just the p-value**, and give the interval.
3. **Be honest about limits**: unequal spreads, observational data, a confounder you cannot measure.
:::

## Recap

| Idea | Takeaway |
|---|---|
| Compare groups | Match the tool to the design: one-sample, two-sample, paired, or ANOVA |
| A significant difference | Does not by itself prove causation |
| Confounding | Something that affects both group membership and the outcome; compare *within* similar units to see its effect |
| Observational comparison | Can be suggestive, but only random assignment can establish cause |
| Practical significance | Look at the size of the effect in real units, not just the p-value |

Next up is Module 4: from *comparing groups* to *modeling relationships* with regression.
