---
title: "Diagnostics: Can This Model Be Trusted?"
subtitle: "Module 4, Part 2 (November 13)"
author: "QBA 5330"
format:
  revealjs:
    theme: simple
    slide-number: true
    footer: "QBA 5330 | Module 4 | Part 2"
    transition: fade
    incremental: true
execute:
  echo: true
  warning: false
  message: false
---

## Recall From Earlier

We trusted every t-test and confidence interval at face value.

**This part: is that trust earned?**

## The LINE Assumptions

- **L**inearity
- **I**ndependence
- **N**ormality
- **E**qual variance

::: {.notes}
Cold-call the room for each letter before revealing. This is recall practice, not new content.
:::

## A Healthy Model

```{r}
#| echo: false
#| message: false
library(tidyverse)
library(olsrr)
ads <- read_csv("data/advertising.csv")
```

```{r}
#| echo: true
#| message: false
model <- lm(sales ~ tv + radio + online, data = ads)
ols_test_breusch_pagan(model)
```


## A Broken Model

```{r}
#| echo: false
set.seed(42)
n <- 150
x3 <- runif(n, 10, 100)
y3 <- 5 + 2*x3 + rnorm(n, 0, x3 * 0.3)
```

```{r}
#| echo: true
model_het <- lm(y3 ~ x3)
ols_test_breusch_pagan(model_het)
```

**Same test. Same code. Opposite conclusion.**

## Activity: Diagnose the Model

Four plots. For each: which LINE assumption (if any) is violated, and what do you do next?

- Plot A: patternless cloud
- Plot B: U-shape
- Plot C: funnel/cone shape
- Plot D: one huge bar in a Cook's distance chart

::: {.notes}
Watch for teams treating Plot D as a LINE violation. It isn't, redirect: LINE is about the model, influence is about individual points.
:::

## Debrief

- Plot A &rarr; no violation, trust it
- Plot B &rarr; linearity, needs a new functional form
- Plot C &rarr; equal variance, standard errors unreliable
- Plot D &rarr; **not a LINE issue** &rarr; one influential point, investigate it

## Before the Break

- Every model today predicted a **number**
- What about yes/no outcomes: will this customer churn?

- **Next: logistic regression, live.**
