---
title: "From Comparing Groups to Modeling Relationships"
subtitle: "Module 4, Part 1 (November 13)"
author: "QBA 5330"
format:
  revealjs:
    theme: simple
    slide-number: true
    footer: "QBA 5330 | Module 4 | Part 1"
    transition: fade
    incremental: true
execute:
  echo: true
  warning: false
  message: false
---

## Where We're Headed Today

- **Part 1:** From comparing groups to modeling relationships
- **Part 2:** Diagnostics: can this model be trusted?
- **Part 3:** Logistic regression and the Module 4 deliverable

::: {.notes}
Module 3 asked "is there a difference?" Today asks "how does x relate to y, and what can we predict?" This is the build-a-model step of the decision cycle.
:::

## Module 3 vs. Today

- **Module 3:** comparing groups (means, proportions), ANOVA, causality
- **Today:** *modeling relationships*, predicting one variable from another

## The Analytics Decision Cycle, Revisited

> Frame the decision &rarr; obtain data &rarr; summarize evidence &rarr; quantify uncertainty &rarr; **build a model** &rarr; recommend an action &rarr; communicate the result

- Everything through Module 3 sat in steps 1&ndash;4
- Today: step 5

## A Familiar Dataset, One Predictor

$$\widehat{\text{sales}} = 10.50 + 0.0527(\text{tv})$$

```{r}
#| echo: true
#| message: false
library(tidyverse)
ads <- read_csv("data/advertising.csv")
model_tv <- lm(sales ~ tv, data = ads)
summary(model_tv)$r.squared
```

## What's Missing?

- The company also spends on **radio** and **online** advertising
- A model using only `tv` might be silently absorbing their effects too

## Adding Predictors, Live

```{r}
#| echo: true
model_full <- lm(sales ~ tv + radio + online, data = ads)
summary(model_full)$coefficients
```

## Watch R-Squared Climb

| Model | R² | Adjusted R² |
|---|---|---|
| tv only | 0.714 | 0.712 |
| tv + radio | 0.856 | 0.854 |
| tv + radio + online | 0.874 | 0.872 |

- Adjusted R² rises **right alongside** raw R² &rarr; real signal, not padding

## The tv Coefficient Actually Moved

- Alone: $b_{tv} = 0.0527$
- With radio and online in the model: $b_{tv} = 0.0512$

- **"Holding other predictors constant" is not just a phrase, it changes the number.**

## Quick Recap: Categorical Predictors

- A dummy coefficient is a comparison **to the reference category**
- Changing the reference category changes what's *reported*, not what's *predicted*

::: {.notes}
Recap only, from @sec-week09-02's store-format example. Don't re-derive it.
:::

## Activity: Interpret This Output (20 minutes)

A snack company surveys 200 households on income, region, and daily servings of fruits/vegetables.

**In groups: what does this model actually tell the company?**

## The Output

```
household_income  -1.277e-06   p = 0.845
regionNorth         0.0342     p = 0.879
regionSouth        -0.3555     p = 0.102
regionWest         -0.1654     p = 0.449

Adjusted R-squared: 0.0001
```

::: {.notes}
Watch for groups calling regionSouth's p = 0.102 "marginally significant." It isn't. This is a deliberately chosen null result.
:::

## Debrief

- What *is* the honest conclusion here?
- Is "we found nothing" still a useful answer for the company?

## Before the Break

- We've been trusting every t-test and confidence interval at face value
- **Hour:** is that trust actually earned?
