---
title: "Logistic Regression, Thresholds, and the Module 4 Deliverable"
subtitle: "Module 4, Part 3 (November 13)"
author: "QBA 5330"
format:
  revealjs:
    theme: simple
    slide-number: true
    footer: "QBA 5330 | Module 4 | Part 3"
    transition: fade
    incremental: true
execute:
  echo: true
  warning: false
  message: false
---

## From Numbers to Yes/No

- Parts 1–2: predicting a **continuous** outcome (sales, revenue)
- Now: predicting a **binary** outcome (churn: yes or no)

## Why Not Just Use Linear Regression?

```{r}
#| echo: true
#| message: false
library(tidyverse)
set.seed(2025)
n <- 300
tenure <- round(runif(n, 1, 60))
logodds_true <- 1.5 - 0.08 * tenure
churned <- rbinom(n, 1, 1 / (1 + exp(-logodds_true)))
customers <- tibble(tenure = tenure, churned = churned)

predict(lm(churned ~ tenure, data = customers), newdata = data.frame(tenure = 100))
```

A "probability" of about $-0.69$. Not usable.

## The Churn Model, Live

```{r}
#| echo: true
model <- glm(churned ~ tenure, data = customers, family = binomial)
coef(model)
exp(coef(model)["tenure"])
```

- Odds ratio $\approx 0.919$: each month of tenure, about **8.1% lower odds** of churn

## One Customer, One Prediction

```{r}
#| echo: true
predict(model, newdata = data.frame(tenure = 12), type = "response")
```

- A 12-month customer: about a **67% chance** of churning

## Activity: Choose Your Threshold (25 minutes)

- Retention offer: **\$50** per customer sent
- Value of a retained customer: **\$600**
- Threshold 0.5: sensitivity 63.2%, specificity 85.1%
- Threshold 0.3: sensitivity 84.9%, specificity 69.1%

- **Which threshold do you recommend, and why?**

## The Numbers to Work With

| Threshold | Flagged | Cost | Caught | Value | Net |
|---|---|---|---|---|---|
| 0.5 | 96 | \$4,800 | 67 | \$40,200 | ? |
| 0.3 | 150 | \$7,500 | 90 | \$54,000 | ? |

::: {.notes}
Let teams compute "Net" themselves ($35,400 and $46,500). Push whichever team picks 0.5 to justify it against these numbers, or to name what would have to change for 0.5 to win.
:::

## Debrief

- With *these* costs, the lower threshold wins
- **What would have to be true for the higher threshold to win instead?**

::: {.notes}
A more expensive offer, a less valuable customer, or an offer that doesn't reliably retain a flagged churner. There is no universally correct threshold.
:::

## The Module 4 Deliverable

- Build a **predictive model** (regression or logistic regression) on a business dataset
- **Diagnose** it
- Translate the result into a **specific managerial recommendation**
- Calculated in R **and** Excel

## Looking Ahead to December 4

Module 4 added **build a model** to the decision cycle.

Next: **forecasting**, predicting a value forward in **time**, culminating in the Ceres Gardening Company case.
