---
title: "Growth vs. Cash Flow: Preparing for the Bank"
subtitle: "Module 5, Part 3 (December 4)"
author: "QBA 5330"
format:
  revealjs:
    theme: simple
    slide-number: true
    footer: "QBA 5330 | Module 5 | Part 3"
    transition: fade
    incremental: true
execute:
  echo: true
  warning: false
  message: false
---

## What the Bank Actually Wants to Know

- Not: "will revenue keep growing?" (Part 2 already answered: yes, comfortably)
- Actually: **growth prospects *and* cash-flow projections**

- **Revenue growth and cash-flow health are not the same question.**

## Two Loan Covenants

- Long-term debt cannot exceed **3.25 &times; EBITDA**
- EBITDA must be at least **2.0 &times; fixed financial charges** (interest + lease rentals)

## Computing Both, Live

```{r}
#| echo: true
ebit_2006 <- 3018; da_2006 <- 669
ebitda_2006 <- ebit_2006 + da_2006
ebitda_2006

interest_2006 <- 658; lease_rentals <- 965
fixed_charges <- interest_2006 + lease_rentals
coverage_ratio <- ebitda_2006 / fixed_charges
coverage_ratio
```

## Two Very Different Pictures

```{r}
#| echo: true
current_debt <- 9000
max_allowed_debt <- 3.25 * ebitda_2006
c(max_allowed = max_allowed_debt, headroom = max_allowed_debt - current_debt)
```

- Debt cap: **~25% headroom**
- Coverage ratio: **2.27x vs. 2.0x minimum — only ~14% cushion**

## Activity: Is Growth Financially Sustainable?

Using the 2007 scenarios from Part 2 and the covenant numbers above:

1. Would rising interest expense (from debt-financed growth) push coverage below 2.0x?
2. Does EBITDA growth loosen the covenants, or could margin compression tighten them?
3. Which covenant should the CEO actually worry about?
4. **Draft a 2–3 sentence recommendation for the bank meeting.**

::: {.notes}
Q1: fixed charges can rise to EBITDA/2.0 ≈ $1,843,500 before breaching, versus $1,623,000 today — room for only about $220,000 more. Q3: the coverage ratio, since it has far less percentage cushion (14%) than the debt cap (25%) — same binding-vs-non-binding logic from week_13_02.
:::

## Report-Outs

**Does your recommendation address the coverage-ratio cushion specifically, or just "pursue the optimistic scenario"?**

## Module 5 Wrap-Up

What this module added:

- The full forecasting toolkit (naive &rarr; regression-based)
- Formal accuracy and forecast-interval concepts
- Scenario analysis — applied to a real financing decision

## Looking Ahead to Tomorrow (12/5)

From **prediction** ("what's likely to happen?") to **prescription** ("what should we do, given real limits on resources?").

**Next: Module 6, the Petro Refinery case.**
