---
title: "Forecasting Ceres's Growth"
subtitle: "Module 5, Part 2 (December 4)"
author: "QBA 5330"
format:
  revealjs:
    theme: simple
    slide-number: true
    footer: "QBA 5330 | Module 5 | Part 2"
    transition: fade
    incremental: true
execute:
  echo: true
  warning: false
  message: false
---

## Ceres's Real Growth

- Revenue: \$24.65M (2002) &rarr; \$42.60M (2006E)
- Year-over-year growth: 8.7%, 9.3%, **19.8%, 21.4%**
- Acceleration coincides with the 2005 launch of **GetCeres™**, a dealer inventory financing program

::: {.notes}
GetCeres offered dealers a 15% discount and 120-day payment terms to encourage them to stock more inventory. It clearly worked, on the revenue side.
:::

## Computing the CAGR

```{r}
#| echo: true
rev <- c(24652, 26797, 29289, 35088, 42597)
names(rev) <- 2002:2006

cagr <- (rev["2006"] / rev["2002"])^(1/4) - 1
cagr
```

**About 14.7% per year, compounded.**

## Base, Optimistic, Pessimistic

```{r}
#| echo: true
base_2006 <- rev["2006"]
base <- as.numeric(cagr)
optimistic <- base + 0.05
pessimistic <- base - 0.10

proj_2007 <- base_2006 * (1 + c(pessimistic, base, optimistic))
round(proj_2007, 0)
```

**A \$6.4 million spread on 2007 revenue alone.**

## All Scenarios Clear Breakeven

- Case breakeven: **~\$30 million** in revenue
- Even the pessimistic scenario: **~\$44.6 million**

- **Revenue growth is not going to be the problem in this case.**

::: {.notes}
Hold this thought — Part 3 is entirely about why "revenue keeps growing" doesn't automatically mean "the company is fine."
:::

## The Problem: Ceres's Reporting Is Annual Only

- 5–6 data points
- No way to detect monthly seasonality or fit a moving average

- **The case itself mentions demand concentrated in just 4–5 weekends each spring/summer.**

## Activity: A Supplementary Monthly Series 

```{r}
#| echo: true
#| message: false
library(tidyverse)
library(fpp3)

set.seed(2025)
months <- seq(as.Date("2004-01-01"), as.Date("2006-12-01"), by = "month")
month_num <- as.integer(format(months, "%m"))
seasonal_shape <- c(0.6,0.7,1.3,1.6,1.5,1.1,0.9,0.8,0.7,0.6,0.5,0.6)
sales <- (700 + 12*(1:length(months))) * seasonal_shape[month_num] + rnorm(length(months), 0, 40)

sell_through <- tibble(month = yearmonth(months), sales = round(sales,1)) |> as_tsibble(index = month)
```

**Fit Holt-Winters. Forecast 6 months. When does the seasonal peak hit?**

::: {.notes}
Peak is March–May, matching the case's own description of spring-weekend-driven demand. Discuss the implication for GetCeres inventory timing: dealers need stock in place *before* the peak, not during it.
:::

## Before the Break

- Revenue is growing. Forecasts confirm it'll likely keep growing.
- But: accounts receivable **\$10.3M &rarr; \$14.5M**; dealer inventory **\$10M &rarr; \$23M** (2005 to 2006E)

- **Does growing revenue automatically mean Ceres's cash position is fine?**
