---
title: "Choosing and Building the Right Forecast"
subtitle: "Module 5, Part 1 (December 4)"
author: "QBA 5330"
format:
  revealjs:
    theme: simple
    slide-number: true
    footer: "QBA 5330 | Module 5 | Part 1"
    transition: fade
    incremental: true
execute:
  echo: true
  warning: false
  message: false
---

## Where We're Headed Today

- **Part 1:** Choosing and building the right forecast
- **Part 2:** Forecasting Ceres's growth
- **Part 3:** Growth vs. cash flow — preparing for the bank

::: {.notes}
Module 4 ended on: "build a model" is now part of the decision cycle; next is forecasting, predicting a value forward in time. Today is that "next."
:::

## Recall: The Full Toolkit

- Naive / seasonal naive
- Moving averages
- Exponential smoothing (SES, Holt's, Holt-Winters)
- Regression-based forecasting (TSLM)

- *(All already in your reading — today applies it, fast.)*

## Live Coding: Back to Back

```{r}
#| echo: true
#| message: false
library(tidyverse)
library(fpp3)

coffee <- read_csv("data/coffee_shop_daily.csv") |> as_tsibble(index = date)
coffee |> autoplot(revenue)
```

## Naive vs. Seasonal Naive

```{r}
#| echo: true
coffee |> model(naive = NAIVE(revenue), snaive = SNAIVE(revenue)) |>
  forecast(h = "1 day") |> hilo(level = 95)
```

## Holt-Winters

```{r}
#| echo: true
coffee |> model(hw = ETS(revenue ~ error("A") + trend("A") + season("A"))) |>
  report()
```

## Activity: Which Method Would You Use?

**No calculation — just diagnosis.**

1. Daily sales: no trend, no pattern, just noise
2. Hourly call volume: strong day-of-week pattern, no trend
3. Monthly revenue: rising trend **and** an end-of-quarter spike
4. Need a same-day estimate, no time to fit anything
5. Want to know exactly how many dollars Friday differs from Monday, accounting for a trend
6. Want to smooth a noisy series to see the trend — not forecasting yet

::: {.notes}
Watch for teams reaching for Holt-Winters or regression by default even in scenario 1, where there's no real trend or seasonality, the simplest tool (SES) is correct there. Complexity has to earn its keep.
:::

## Debrief

- Scenario 3: what would you actually fit?

::: {.notes}
Regression-based forecasting (TSLM) with trend() plus a quarter-end indicator, or Holt-Winters, is the right answer; walk through why.
:::

## Later: A Real Company

- Ceres Gardening Company's CEO has two things on his to-do list:
  - A motivational speech for the sales team
  - A meeting with the **bank** about growth and cash flow

- **We're helping with the second one.**

## Before the Break

- Ceres's own financial reporting is **annual only** — 5–6 data points
- Nowhere near enough for moving averages or seasonality

- **Next: a supplementary monthly series, then back to Ceres's real numbers.**
