# Instructor Guide — Part 1 (12/5, Hour 1): Building Petro Refinery's Model {#sec-mod6-part1-guide}

::: {.callout-note collapse="false" title="How to use this guide"}
This is a facilitation guide for the instructor, not student-facing reading. It pairs with the slide deck `mod6_part1_slides.qmd`. Students arrive having read the mechanics of linear optimization (@sec-week13-01) and binding constraints, slack, and sensitivity (@sec-week13-02) using a furniture-company example. This hour builds the real Petro Refinery LLC model from scratch in Excel Solver, following the exact setup steps from @sec-week13-01, then verifies and interprets the result.
:::

## Session Objectives

By the end of this period, students should be able to:

1. Translate Petro Refinery's product-mix decision into decision variables, an objective function, and constraints, in both words and math.
2. Build and solve the model in Excel Solver, reproducing the case's own optimal answer.
3. Correctly identify which of Petro's three resource constraints are binding and which have slack.

## Timing at a Glance

| Time | Segment | Format |
|---|---|---|
| 0:00–0:10 | Welcome and reconnecting to the reading | Instructor-led |
| 0:10–0:20 | Setting up Petro's model in words and math | Instructor-led |
| 0:20–0:45 | Live build: Petro in Excel Solver | Live coding (Excel) |
| 0:45–0:55 | Verifying in R; identifying binding constraints | Live coding (R) |
| 0:55–1:00 | Debrief and bridge to Part 2 | Instructor-led |

## Detailed Run of Show

### Welcome and Reconnecting to the Reading (10 min)

Recall the furniture-company example from @sec-week13-01 and @sec-week13-02 (tables and chairs, wood and labor and finishing constraints). Restate the parallel explicitly: Petro Refinery's problem has exactly the same *shape*, two decision variables, a linear objective, a handful of resource constraints, just a different business (an oil refinery deciding how much light and heavy crude to process, rather than a furniture shop deciding how many tables and chairs to build).

### Setting Up Petro's Model in Words and Math (10 min)

Walk the room through the setup exactly as the case presents it:

- **Decision variables**: $X_1$ = barrels of light crude to process (in thousands) each week; $X_2$ = barrels of heavy crude to process (in thousands) each week.
- **Objective**: light crude contributes \$10 per barrel toward fixed costs, heavy crude \$12.5 per barrel. Maximize $10X_1 + 12.5X_2$ (total weekly contribution, in \$ thousands).
- **Constraints**: three shared pieces of refinery equipment, distillation, conversion, and blending, each with limited weekly hours, and each barrel of light or heavy crude uses a different amount of each:
$$
\begin{align*}
\text{Distillation:} \quad & 2X_1 + 2X_2 \leq 3{,}000 \\
\text{Conversion:} \quad & 2X_1 + 3X_2 \leq 2{,}400 \\
\text{Blending:} \quad & 3X_1 + 2.5X_2 \leq 2{,}400 \\
& X_1, X_2 \geq 0
\end{align*}
$$

### Live Build: Petro in Excel Solver (25 min)

Reproduce @sec-week13-01's Solver setup steps live, on Petro's actual numbers:

1. Decision variable cells for $X_1$ and $X_2$; an objective cell computing $10X_1 + 12.5X_2$.
2. Three constraint-usage cells, one per resource, each a formula like the objective's (coefficients times decision variables).
3. Open Solver: **Set Objective** to the contribution cell (**Max**), **By Changing Variable Cells** to $X_1$ and $X_2$, **Subject to the Constraints** for all three resource cells, **Simplex LP** as the solving method, **Make Unconstrained Variables Non-Negative** checked.
4. **Solve.**

Solver returns $X_1=300$ (thousand barrels of light crude), $X_2=600$ (thousand barrels of heavy crude), for a maximum weekly contribution of **\$10,500,000**.

::: {.callout-note collapse="false" title="A natural student question to address directly"}
Heavy crude has the higher per-barrel margin (\$12.5 vs. \$10), so why doesn't Petro just process heavy crude exclusively? Answer: because heavy crude also uses relatively *more* of the conversion and blending resources per barrel of contribution earned. The optimal mix reflects a genuine trade-off across three shared, limited resources, not just a ranking of which product is individually more profitable. This is exactly why the LP needs all three constraints, not just the objective, to find the real answer.
:::

### Verifying in R; Identifying Binding Constraints (10 min)

```{r}
#| echo: true
library(lpSolve)

obj <- c(10, 12.5)
mat <- matrix(c(2, 2,
                 2, 3,
                 3, 2.5), nrow = 3, byrow = TRUE)
rhs <- c(3000, 2400, 2400)
dir <- c("<=", "<=", "<=")

sol <- lp("max", obj, mat, dir, rhs)
sol$solution
sol$objval
```
Confirm this matches Solver exactly. Then compute resource usage and slack live:
```{r}
#| echo: true
used <- mat %*% sol$solution
data.frame(resource = c("Distillation", "Conversion", "Blending"),
           available = rhs, used = as.numeric(used), slack = rhs - as.numeric(used))
```
**Conversion and blending are both fully used** (binding, zero slack); **distillation has 1,200 hours of slack**, only 60% of its available 3,000 hours are actually needed at the optimal mix. Distillation capacity is not what's limiting Petro's profit right now.

### Debrief and Bridge to Part 2 (5 min)

Ask the room: "if Petro's management wanted to increase profit further, would buying more distillation capacity help?" Let the (correct) answer, no, since it's non-binding, land before revealing why formally. Bridge: "Part 2 puts a precise dollar value on exactly this kind of question, for every resource, using the Sensitivity Report."

## Recap for the Instructor

| **Keyword** | **Where it was taught** |
|---|---|
| LP building blocks, Excel Solver setup | @sec-week13-01 |
| Binding vs. non-binding constraints, slack | @sec-week13-02 |
| Petro Refinery's actual model and solution | This session (drawn from the case) |
