---
title: "Building Petro Refinery's Model"
subtitle: "Module 6, Part 1 (December 5)"
author: "QBA 5330"
format:
  revealjs:
    theme: simple
    slide-number: true
    footer: "QBA 5330 | Module 6 | Part 1"
    transition: fade
    incremental: true
execute:
  echo: true
  warning: false
  message: false
---

## Where We're Headed Today

- **Part 1:** Building Petro Refinery's model
- **Part 2:** The value of an extra hour — sensitivity analysis
- **Part 3:** Executive recommendation and course wrap-up

::: {.notes}
Same shape as yesterday's furniture example: two decision variables, a linear objective, a few resource constraints. Different business.
:::

## Petro Refinery LLC

A crude oil refinery deciding how much **light** and **heavy** crude to process each week.

- $X_1$ = barrels of light crude (thousands)
- $X_2$ = barrels of heavy crude (thousands)

## The Model

$$
\text{Maximize } 10X_1 + 12.5X_2
$$

$$
\begin{align*}
\text{Distillation:} \quad & 2X_1 + 2X_2 \leq 3{,}000 \\
\text{Conversion:} \quad & 2X_1 + 3X_2 \leq 2{,}400 \\
\text{Blending:} \quad & 3X_1 + 2.5X_2 \leq 2{,}400
\end{align*}
$$

## Why Not Just Make Heavy Crude?

- Heavy crude's margin is higher (\$12.5 vs. \$10)
- But heavy crude also uses **more** conversion and blending per barrel

- **The optimal mix is a genuine trade-off, not a ranking.**

## Live Build: Excel Solver

1. Decision variable cells + objective cell
2. One usage cell per constraint
3. Solver: Set Objective (Max), By Changing Cells, Subject to Constraints
4. Simplex LP, Assume Non-Negative
5. **Solve**

## The Answer

**300,000 barrels of light crude. 600,000 barrels of heavy crude.**

**\$10,500,000 per week in contribution.**

## Verifying in R

```{r}
#| echo: true
library(lpSolve)
obj <- c(10, 12.5)
mat <- matrix(c(2,2, 2,3, 3,2.5), nrow = 3, byrow = TRUE)
rhs <- c(3000, 2400, 2400)
sol <- lp("max", obj, mat, c("<=","<=","<="), rhs)
sol$solution
sol$objval
```

## Binding or Not?

```{r}
#| echo: true
used <- mat %*% sol$solution
data.frame(resource = c("Distillation","Conversion","Blending"),
           available = rhs, used = as.numeric(used), slack = rhs - as.numeric(used))
```

- Conversion & blending: **fully used**
- Distillation: **1,200 hours of slack** (only 60% used)

## Before the Break

**Would buying more distillation capacity increase Petro's profit?**

::: {.notes}
No — it's non-binding. Let this land before Part 2 formalizes exactly why with shadow prices.
:::

**Next: putting a precise dollar value on every resource.**
