# Week 11: Introduction to Logistic Regression {#sec-week11-01}

## Why This Matters

Every regression model in this course so far has predicted a **quantitative** response, sales, revenue, a rating. Plenty of the most important business questions have a **binary** answer instead: will this customer churn or stay? Will this loan default or not? Will this lead convert or not? The response variable in each case only takes two values, usually coded 0 and 1, and predicting it calls for a different tool: **logistic regression**, the beginning of this course's coverage of classification.

## Why Not Just Use Linear Regression?

It's tempting to just fit an ordinary least-squares line to a 0/1 outcome. It runs without error, but the result doesn't behave like a probability.

::: {.example title="Example 11.1: What goes wrong with a linear fit"}
A subscription company tracks 300 customers' account `tenure` (in months) and whether each `churned` (1) or stayed (0) over the following quarter.

```{r}
#| echo: true
#| message: false
library(tidyverse)
set.seed(2025)
n <- 300
tenure <- round(runif(n, 1, 60))

#simulate churn probability that genuinely declines with tenure
logodds_true <- 1.5 - 0.08 * tenure
p_true <- 1 / (1 + exp(-logodds_true))
churned <- rbinom(n, 1, p_true)
customers <- tibble(tenure = tenure, churned = churned)

model_linear <- lm(churned ~ tenure, data = customers)
range(fitted(model_linear))
predict(model_linear, newdata = data.frame(tenure = 100))
```
The fitted values already dip slightly **below 0** within the observed data, and predicting a customer with 100 months of tenure gives about $-0.69$, a "probability" that's meaningless. A straight line has no way to know it should flatten out near 0 and 1; it just keeps going. There's also a structural problem the range issue doesn't even capture: with a binary response, the residuals can't be normally distributed or have constant variance (they only take two possible values at each $x$), so the LINE assumptions behind every t-test and confidence interval from @sec-week10-01 don't hold to begin with.
:::

## Probability, Odds, and Log-Odds

Logistic regression solves this by modeling a *transformed* version of probability that isn't boxed in between 0 and 1.

- **Probability** ($p$): between 0 and 1, the chance the event (churn, default, conversion) happens.
- **Odds**: $\dfrac{p}{1-p}$, the ratio of the event happening to it not happening. Odds range from 0 (impossible) to $\infty$ (certain), and are common business language already ("2-to-1 odds").
- **Log-odds** (the **logit**): $\ln\!\left(\dfrac{p}{1-p}\right)$. Taking the log stretches odds' $(0,\infty)$ range out to the entire real number line, $(-\infty,\infty)$, exactly the range a linear model's right-hand side can already produce.

| $p$ | Odds ($p/(1-p)$) | Log-odds |
|---|---|---|
| 0.10 | 0.111 | $-2.20$ |
| 0.50 | 1.000 | $0.00$ |
| 0.90 | 9.000 | $2.20$ |

Notice $p=0.5$ is the natural center: odds of exactly 1 (even odds), log-odds of exactly 0.

## The Logistic Regression Model

Logistic regression models the log-odds as a linear function of the predictors, exactly the right-hand side we've used all module:
$$
\ln\!\left(\frac{p}{1-p}\right) = b_0 + b_1 x
$$
Solving for $p$ gives the **logistic function**, an S-shaped (sigmoid) curve that is mathematically guaranteed to stay between 0 and 1 for any input:
$$
p = \frac{1}{1+e^{-(b_0+b_1x)}}
$$
This is the fix for Example 11.1's problem: no matter how extreme $x$ gets, $p$ can get arbitrarily close to 0 or 1, but never cross those boundaries.

::: {.example title="Example 11.2: Fitting the logistic regression"}
```{r}
#| echo: true
model_logit <- glm(churned ~ tenure, data = customers, family = binomial)
summary(model_logit)
```
The fitted model, on the log-odds scale, is
$$
\ln\!\left(\frac{\hat{p}}{1-\hat{p}}\right) = 1.720 - 0.0846\,(\text{tenure})
$$
The `tenure` coefficient is negative and statistically significant ($p<0.0001$): longer-tenured customers have lower log-odds of churning. That's a useful sign check, but "log-odds" isn't a natural unit for a business conversation; @sec-week11-02 translates this coefficient into something far easier to communicate, an **odds ratio**.
:::

## Predicted Probabilities

Even before translating the coefficient itself, the fitted model can convert any specific `tenure` value into a predicted probability using the logistic function above.

::: {.example title="Example 11.3: The churn curve"}
```{r}
#| echo: true
predict(model_logit, newdata = data.frame(tenure = c(1, 12, 36, 60)), type = "response")
```
Predicted churn probability falls from about **83.7%** at 1 month of tenure to **66.9%** at one year, **21.0%** at three years, and just **3.4%** at five years, a steep, plausible decline consistent with the idea that customers who have stuck around are increasingly unlikely to leave.
```{r}
#| echo: false
customers |>
  mutate(fitted_p = predict(model_logit, type = "response")) |>
  arrange(tenure) |>
  ggplot(aes(x = tenure)) +
  geom_jitter(aes(y = churned), height = 0.03, alpha = 0.3, color = "#2c7fb8") +
  geom_line(aes(y = fitted_p), color = "#e34a33", linewidth = 1) +
  labs(title = "Predicted probability of churn vs. tenure",
       subtitle = "Points: actual outcomes (jittered). Curve: fitted logistic model.",
       x = "Tenure (months)", y = "Probability of churn")
```
The S-curve shape is exactly what the logistic function guarantees: steep in the middle range of tenure, where a customer's fate is genuinely uncertain, and flattening out toward 0 and 1 at the extremes, where the outcome is increasingly predictable.
:::

## How Logistic Regression Is Actually Fit

Linear regression's coefficients came from **least squares**, minimizing the sum of squared residuals. That criterion doesn't transfer cleanly to a 0/1 outcome. Instead, logistic regression uses **maximum likelihood estimation**: it searches for the coefficients $b_0, b_1$ that make the actually observed pattern of 0s and 1s in the data as probable as possible, given the logistic model's assumed shape. There's no simple closed-form formula the way there was for simple linear regression's slope and intercept in @sec-week08-01; software finds these coefficients through an iterative numerical search (the `summary()` output's "Fisher Scoring iterations" line in Example 11.2 reflects that search). The practical takeaway: rely on software for the fitting itself, and focus your own effort on interpreting what comes out.

## Computing Logistic Regression in R and Excel

::: {.example title="Example 11.4: Logistic regression in R and Excel"}
In R, `glm()` with `family = binomial` fits a logistic regression exactly as `lm()` fits a linear one, shown throughout this section.

::: {.callout-note collapse="false" title="Excel has no built-in logistic regression tool"}
Unlike ordinary regression, the Data Analysis ToolPak has no logistic regression option. Fitting one in Excel requires setting up the log-likelihood function by hand and maximizing it with **Solver**, considerably more manual work than any other technique in this course. For any real logistic regression analysis, this course recommends fitting the model in R with `glm()`, even if the surrounding exploratory work is done in Excel.
:::
:::

## Recap

| **Keyword** | **Definition** |
|---|---|
| **Logistic regression** | A regression model for a binary (0/1) response, modeling the log-odds of the outcome as a linear function of the predictors. |
| **Odds** | $p/(1-p)$; the ratio of an event happening to it not happening. |
| **Log-odds (logit)** | $\ln(p/(1-p))$; unbounded, exactly what a linear model's right-hand side can produce. |
| **Logistic function** | $p = 1/(1+e^{-(b_0+b_1x)})$; converts log-odds back to a probability, mathematically bounded between 0 and 1. |
| **Maximum likelihood estimation** | The method used to fit logistic regression; finds the coefficients that make the observed 0/1 outcomes most probable under the model. |

## Check Your Understanding

::: {.callout-note collapse="false" title="Problems"}
1. A bank tries fitting an ordinary linear regression to predict loan default (1) vs. no default (0) from credit score. Describe two distinct problems with this approach, beyond just "it's not the standard method."

2. If $p=0.20$, compute the odds and the log-odds. If $p=0.80$, compute the odds and log-odds. What do you notice about the relationship between these two results?

3. A logistic regression predicting customer conversion from `minutes_on_site` gives $b_1 = 0.15$ (on the log-odds scale). Without computing an exact probability, explain what the sign of this coefficient tells you in plain business language.

4. Using the logistic function $p = 1/(1+e^{-(b_0+b_1x)})$, explain why $p$ can never actually reach exactly 0 or exactly 1, no matter how extreme $x$ becomes.

5. Explain, in your own words, why logistic regression can't be fit using the least-squares method from earlier in this module, and what method is used instead.
:::

::: {.callout-tip collapse="true" title="Solutions"}
1. First, the model's predicted "probabilities" can fall below 0 or above 1 for extreme credit scores, values that make no sense as probabilities. Second, since the response only takes values 0 and 1, the residuals cannot be normally distributed or have constant variance across fitted values, violating the LINE assumptions (@sec-week10-01) that the model's standard errors, p-values, and confidence intervals all depend on.

2. At $p=0.20$: odds $=0.20/0.80=0.25$, log-odds $=\ln(0.25)\approx -1.386$. At $p=0.80$: odds $=0.80/0.20=4.00$, log-odds $=\ln(4.00)\approx 1.386$. The log-odds are exact negatives of each other ($-1.386$ vs. $1.386$), reflecting that $p=0.20$ and $p=0.80$ are symmetric around $p=0.5$ (log-odds of 0).

3. A positive coefficient means more minutes spent on the site is associated with **higher** log-odds (and therefore higher probability) of conversion, more engaged visitors, in terms of time spent, are more likely to convert, holding nothing else in the model constant.

4. As $x\to\infty$ (with $b_1>0$), $e^{-(b_0+b_1x)} \to 0$, so $p \to 1/(1+0) = 1$, but never exactly reaches it for any finite $x$. As $x\to -\infty$, $e^{-(b_0+b_1x)} \to \infty$, so $p \to 1/\infty = 0$, again only approached, never reached. The function gets arbitrarily close to 0 and 1 but is mathematically guaranteed to stay strictly between them for any finite input.

5. Least squares minimizes the sum of squared residuals, a criterion built around a *quantitative* response where "closeness" between predicted and actual values is measured on a continuous scale. With a 0/1 response, that geometric idea doesn't transfer cleanly, and it also isn't the criterion that gives logistic regression its bounded-probability guarantee. Instead, logistic regression uses **maximum likelihood estimation**, which chooses the coefficients that make the actual observed pattern of 0s and 1s as probable as possible under the assumed logistic model, found through an iterative numerical search rather than a direct formula.
:::
