# Week 3: The Uniform Distribution {#sec-week03-02}

## Why This Matters

The binomial distribution counts things, a whole number of successes out of a fixed number of trials. Many business quantities aren't counts at all: a delivery time, a wait time, a dollar amount. These are **continuous random variables**, which can take any value within a range rather than a countable list of values. We'll spend the next two sections on continuous distributions, starting with the simplest one, the **uniform distribution**, before moving to the single most important one, the normal distribution. The uniform distribution isn't just a warm-up exercise, either. It's the engine behind every random number your computer generates, which makes it the natural starting point for simulation, something we'll lean on again later in this course.

## From Discrete to Continuous

A continuous random variable can take any value in an interval: a delivery could arrive at 2:00pm, 2:01pm, or 2:00:47.3pm. Because there are infinitely many possible values, it doesn't make sense to ask for the probability of one *exact* value, that probability is always 0. Instead, we ask for the probability that the variable falls in some **range**, and we get that probability from the *area under a curve* describing the distribution, rather than from a table like we used for discrete random variables.

For most continuous distributions, computing that area requires calculus. The uniform distribution is the happy exception: its curve is just a flat rectangle, so the area is nothing more than base times height arithmetic, not calculus.

## The Uniform Distribution

A continuous random variable $X$ follows a **Uniform$(c,d)$** distribution if every value between $c$ and $d$ is equally likely, and no value outside that range is possible. Its density is a flat, horizontal line of height $\dfrac{1}{d-c}$ across the interval $(c,d)$&mdash;a rectangle, chosen so that its total area equals 1 (as it must, since total probability is always 1).

::: {.example title="Example 3.6: A same-day delivery window"}
A retailer promises same-day delivery "sometime between 9am and 5pm," and internally, the delivery time is equally likely to be any moment in that 8-hour window. Let $T$ = hours after 9am that the delivery arrives, so $T \sim \text{Uniform}(0, 8)$.

```{r}
#| echo: false
library(ggplot2)

c_val <- 0
d_val <- 8

df <- data.frame(x = c(c_val, d_val), y = c(1/(d_val - c_val), 1/(d_val - c_val)))

ggplot(df, aes(x = x, y = y)) +
  geom_line(color = "#2c7fb8", linewidth = 1) +
  geom_segment(aes(x = c_val, xend = c_val, y = 0, yend = 1/(d_val - c_val)), linetype = 2) +
  geom_segment(aes(x = d_val, xend = d_val, y = 0, yend = 1/(d_val - c_val)), linetype = 2) +
  scale_x_continuous(breaks = c(0, 8), labels = c("9am", "5pm")) +
  ylim(0, 0.2) +
  labs(title = "Density of delivery time, T ~ Uniform(0, 8)", x = "Time", y = "Density") +
  theme_minimal()
```
:::

Because the density is a flat rectangle, the probability that $X$ falls between any two points $a$ and $b$ (with $c \le a < b \le d$) is simply the width of that interval divided by the total width:
$$
P(a < X < b) = \frac{b-a}{d-c}
$$

::: {.example title="Example 3.7: Probability of a noon-to-1pm delivery"}
Continuing Example 3.6, what's the probability a delivery arrives between noon and 1pm? Noon is 3 hours after 9am and 1pm is 4 hours after 9am, so $a=3$, $b=4$, and the full window is $(c,d)=(0,8)$:
$$
P(3 < T < 4) = \frac{4-3}{8-0} = \frac{1}{8} = 0.125
$$
Notice that this only depends on the *length* of the target interval (1 hour) relative to the *length* of the whole window (8 hours)&mdash;not on where in the window it falls. A 1-hour window from 2pm to 3pm would give exactly the same probability.
:::

The mean and standard deviation of a Uniform$(c,d)$ distribution have simple closed forms:
$$
\mu = \frac{c+d}{2} \qquad\qquad \sigma = \frac{d-c}{\sqrt{12}}
$$

::: {.example title="Example 3.8: Mean and SD of the delivery window"}
For $T \sim \text{Uniform}(0,8)$ from Example 3.6:
$$
\begin{align*}
\mu &= \frac{0+8}{2} = 4 \text{ hours after 9am (i.e., 1pm)}\\\\ \sigma &= \frac{8-0}{\sqrt{12}} \approx 2.31 \text{ hours}
\end{align*}
$$
The "expected" delivery time is 1pm, right in the middle of the window, which makes sense given every moment in the window is equally likely.
:::

## The Uniform Distribution and Simulation

Here's why the uniform distribution matters beyond describing delivery windows: it's the building block behind essentially every random simulation you'll run in this course. When Excel's `RAND()` or R's `runif()` generates a "random number," it's drawing from a **Uniform(0,1)** distribution&mdash;and every other kind of randomness a computer simulates (random assignment to a treatment group, a Monte Carlo scenario, a shuffled sample) is built by transforming draws from that same Uniform(0,1) starting point.

::: {.example title="Example 3.9: Uniform probabilities and simulation in R and Excel"}
To compute the exact probability from Example 3.7 in R, use `punif()`:
```{r}
#| echo: true
punif(4, min = 0, max = 8) - punif(3, min = 0, max = 8)
```
This matches our by-hand answer of 0.125: it's the cumulative probability up to 4 hours, minus the cumulative probability up to 3 hours.

To *simulate* 10,000 random delivery times and check that they behave the way theory predicts:
```{r}
#| echo: true
#| message: false
library(tidyverse)

set.seed(2025)
sim_deliveries <- runif(10000, min = 0, max = 8)
mean(sim_deliveries)                              # should be close to 4
mean(sim_deliveries > 3 & sim_deliveries < 4)      # should be close to 0.125
```

In Excel, `=RAND()*8` generates a single simulated delivery time on $(0,8)$, and dragging that formula down a column simulates many deliveries at once&mdash;the same logic we'll use later in the course to build simulation-based what-if analyses.
:::

## Recap

| **Keyword** | **Definition** |
|---|---|
| **Continuous random variable** | A random variable that can take any value within an interval; the probability of any single exact value is 0. |
| **Uniform distribution** | A continuous distribution on $(c,d)$ where every value in the interval is equally likely; density is constant at $1/(d-c)$. |
| **Uniform probability rule** | $P(a<X<b) = (b-a)/(d-c)$&mdash;the ratio of interval lengths. |
| **Uniform mean and SD** | $\mu = (c+d)/2$, $\sigma = (d-c)/\sqrt{12}$. |

## Check Your Understanding

::: {.callout-note collapse="false" title="Problems"}
1. A ride-share app tells riders their driver will arrive "in 2 to 10 minutes," and arrival time within that window is equally likely at any point. Let $X$ = arrival time in minutes. What distribution does $X$ follow, and what are its mean and standard deviation?

2. Using the distribution from Problem 1, what is the probability the driver arrives within the first 4 minutes (between 2 and 4 minutes)?

3. A factory's automated filling machine dispenses between 495 mL and 505 mL into a bottle, uniformly at random. What's the probability a randomly selected bottle contains between 498 and 502 mL?

4. Explain, without doing any calculation, why $P(2 < X < 3)$ and $P(7 < X < 8)$ are equal for the ride-share example in Problem 1.
:::

::: {.callout-tip collapse="true" title="Solutions"}
1. $X \sim \text{Uniform}(2, 10)$. $\mu = (2+10)/2 = 6$ minutes. $\sigma = (10-2)/\sqrt{12} \approx 2.31$ minutes.

2. $P(2 < X < 4) = \dfrac{4-2}{10-2} = \dfrac{2}{8} = 0.25$.

3. Let $Y \sim \text{Uniform}(495, 505)$. $P(498 < Y < 502) = \dfrac{502-498}{505-495} = \dfrac{4}{10} = 0.4$.

4. For a uniform distribution, the probability of an interval depends only on the interval's *length*, not its location within $(c,d)$. Both $(2,3)$ and $(7,8)$ have length 1, and both lie within the Uniform$(2,10)$ range, so they have the same probability regardless of where they fall in the window.
:::
