# Week 2: Random Variables {#sec-week02-02}

## Why This Matters

Last section, we worked with probabilities of events (a transaction is fraudulent, a call converts, a unit is defective). To go further to compute an expected profit, describe how a count of monthly defects varies, or eventually build the probability distributions behind confidence intervals, we need a way to attach a **number** to an uncertain outcome and describe how likely each possible number is. That's what a **random variable** does, and it's the building block for everything else in this module.

## What Is a Random Variable?

A **random variable** is a rule that assigns a numerical value to each outcome of a random process. Formally, it's a function whose domain is the sample space and whose range is a set of numbers, but informally, it's just the "number we care about" from an uncertain situation.

::: {.example title="Example 2.11: Turning outcomes into numbers"}
An online retailer wants to study how many items customers add to their cart during a single visit. The underlying outcome (a customer's entire browsing session) is complicated: pages viewed, time spent, items considered and rejected. The random variable
$$
X = \text{number of items added to the cart}
$$
distills all of that complexity down to a single number. Before we observe a given customer's session, $X$ could be 0, 1, 2, 3, or more. Once the session ends, we observe one specific value&mdash;say $x=2$, which is called a **realization** of $X$.
:::

Random variables come in two broad types:

- A **discrete random variable** takes a countable set of values, usually arising from counting: number of items in a cart, number of support tickets filed today, number of defective units in a shipment.
- A **continuous random variable** can take any value in an interval: the time until a support ticket is resolved, the exact weight of a shipped package. We'll study continuous random variables in an upcoming week; for now, our focus is discrete random variables.

::: {.callout-note collapse="false" title="A random variable describes a process, not a single dataset"}
Before a call center's shift starts, "number of complaints received today" is unknown and could take many values - that's the random variable. Once the shift ends and we've counted 7 complaints, that 7 is one realization, not the random variable itself. Run the same shift again next week and you'll likely observe a different number, even though nothing about the underlying process changed. This is why we use an uppercase letter ($X$) for the random variable and a lowercase letter ($x$) for an observed realization.
:::

## Probability Distributions

Once we have a discrete random variable, the next step is describing how likely each of its possible values is. A **probability distribution** assigns a probability $P(X=x)$ to every possible value $x$. Any valid probability distribution must satisfy two conditions:

1. **Non-negativity:** every probability satisfies $0 \le P(X=x) \le 1$.
2. **Sums to one:** $\displaystyle\sum_x P(X=x) = 1$, something in the sample space has to happen.

::: {.example title="Example 2.12: Distribution of items per cart"}
Based on a large sample of past sessions, the retailer estimates the following distribution for $X$ = number of items added to the cart:

| $x$ | 0 | 1 | 2 | 3 or more |
|---|---|---|---|---|
| $P(X=x)$ | 0.45 | 0.30 | 0.15 | 0.10 |

Every probability is between 0 and 1, and they sum to $0.45+0.30+0.15+0.10=1$, so this is a valid distribution. It tells the retailer that, over many sessions, about 45% end with an empty cart, 30% with exactly one item, and so on.

```{r}
#| echo: true
#| message: false
library(tidyverse)

cart_dist <- tibble(
  items = c("0", "1", "2", "3+"),
  prob  = c(0.45, 0.30, 0.15, 0.10)
)

ggplot(cart_dist, aes(x = items, y = prob)) +
  geom_col(fill = "#2c7fb8") +
  labs(title = "Distribution of items added to cart", x = "Items added", y = "Probability")
```
:::

::: {.callout-note collapse="false" title="Checking whether a proposed distribution is valid"}
Given a proposed distribution, check both conditions before using it: are all the probabilities between 0 and 1, and do they sum to 1 (allowing for small rounding)? A distribution that fails either check (e.g., includes a negative probability, or sums to 1.1) cannot be a real probability distribution and needs to be corrected first.
:::

## Expected Value and Variance of a Random Variable

Just as we describe a *sample* with statistics like $\bar{x}$ and $s$, we describe a random variable (really, an entire population or process) with **parameters**, typically written with Greek letters ($\mu$, $\sigma$, $\sigma^2$).

### Expected Value

The **expected value** (or mean) of a discrete random variable is a probability-weighted average of its possible values:
$$
\mu = E(X) = \sum_x x \cdot P(X=x)
$$
You can think of $E(X)$ as the long-run average of $X$ if the underlying process were repeated many, many times&mdash;it is not necessarily a value $X$ can actually take on any single trial.

::: {.example title="Example 2.13: Should we sell an extended warranty?"}
An electronics retailer sells a $1,200 laptop and is deciding whether to offer a one-year extended warranty for $80. Based on historical claims data, there's a 6% chance a customer will file a claim during that year, and the average repair cost on a valid claim is $650. Let $X$ = profit to the retailer per warranty sold:

- If no claim is filed: profit is $\$80$.
- If a claim is filed: profit is $\$80 - \$650 = -\$570$.

$$
\begin{align*}
E(X) &= (80)(0.94) + (-570)(0.06)\\
&= 75.20 - 34.20\\
&= 41.00
\end{align*}
$$

On average, the retailer expects to make **$41 per warranty sold**, even though any individual warranty either nets $80 (no claim) or loses $570 (a claim). This is exactly why insurance and warranty products are priced this way: the expected value is positive for the seller even though any single outcome could be a loss.
:::

::: {.callout-note collapse="false" title="Expected value is a long-run average, not a prediction"}
$E(X) = \$41$ does **not** mean the retailer makes $41 on every warranty&mdash;no single customer's outcome will actually equal $41. It means that if the retailer sells this warranty to thousands of similar customers, the *average* profit per warranty will be close to $41. Expected value is a planning tool for volume, not a forecast for one transaction.
:::

::: {.example title="Example 2.14: The Florida Lottery"}
Expected value doesn't have to come out positive, it's just as useful for showing when a "bet" is a bad one, on average. The Florida Lottery runs two popular daily games, Pick 3 and Pick 4.

In **Pick 3**, players pay \$1 to select three numbers in order, each ranging from 0 to 9. If all three numbers match the order of the numbers drawn, the player wins \$500. The probability of winning Pick 3 is 0.001.

Let $X$ be the amount of money a player nets from a single Pick 3 ticket. The possible values are:

- If the player loses: $X = -\$1$
- If the player wins: $X = \$500 - \$1 = \$499$

$$
\begin{align*}
E(X) &= (-1)(0.999) + (499)(0.001)\\
&= -0.999 + 0.499\\
&= -0.50
\end{align*}
$$

**Pick 4** is similar, but players must match four numbers in order for a \$5,000 payout, and the probability of winning is only 0.0001. Let $X$ be the net amount from a single Pick 4 ticket:

- If the player loses: $X = -\$1$
- If the player wins: $X = \$5{,}000 - \$1 = \$4{,}999$

$$
\begin{align*}
E(X) &= (-1)(0.9999) + (4999)(0.0001)\\
&= -0.9999 + 0.4999\\
&= -0.50
\end{align*}
$$

Both games have the same expected value: **-$0.50 per ticket**. Despite very different odds and payouts, a player loses 50 cents on average for every dollar wagered, on either game. This is exactly how lottery games (and casinos) are designed to work: the expected value to the *player* is reliably negative, which is what makes the expected value to the *operator* reliably positive, the same logic as the warranty example above, just from the other side of the transaction.
:::

### Variance and Standard Deviation

The **variance** of $X$ measures how spread out its values are around the mean:
$$
\sigma^2 = \sum_x (x-\mu)^2 \cdot P(X=x)
$$
The **standard deviation**, $\sigma = \sqrt{\sigma^2}$, is in the same units as $X$ and describes a typical distance from the mean.

::: {.example title="Example 2.15: Variability in items returned per order"}
A retailer's data shows the number of items returned per order, $Y$, follows this distribution:

| $y$ | 0 | 1 | 2 | 3 |
|---|---|---|---|---|
| $P(Y=y)$ | 0.70 | 0.20 | 0.07 | 0.03 |

The expected number of items returned per order is
$$
\mu_Y = 0(0.70) + 1(0.20) + 2(0.07) + 3(0.03) = 0.43
$$

To find the variance, compute $(y-\mu)^2$ for each value, weight by its probability, and sum:

| $y$ | $(y - 0.43)^2$ | $P(Y=y)$ | Contribution |
|---|---|---|---|
| 0 | 0.1849 | 0.70 | 0.12943 |
| 1 | 0.3249 | 0.20 | 0.06498 |
| 2 | 2.4649 | 0.07 | 0.17254 |
| 3 | 6.6049 | 0.03 | 0.19815 |

$$
\sigma_Y^2 = 0.12943 + 0.06498 + 0.17254 + 0.19815 \approx 0.5651
$$
$$
\sigma_Y = \sqrt{0.5651} \approx 0.752
$$

An average order has 0.43 returned items, with a typical order deviating from that average by about 0.75 items. Note that neither the mean nor the standard deviation has to equal a value $Y$ can actually take, you cannot return 0.43 items, they summarize the *distribution*, not any one order.
:::

::: {.example title="Example 2.16: Computing expected value and variance in R"}
```{r}
#| echo: true
returns <- tibble(
  y = c(0, 1, 2, 3),
  prob = c(0.70, 0.20, 0.07, 0.03)
)

mu <- sum(returns$y * returns$prob)
var <- sum((returns$y - mu)^2 * returns$prob)

mu
var
sqrt(var)
```
In Excel, the same calculation uses `SUMPRODUCT()`: `=SUMPRODUCT(y_range, prob_range)` for the mean, and a helper column for $(y-\mu)^2$ multiplied into `SUMPRODUCT()` again for the variance.
:::

## Recap

| **Keyword** | **Definition** |
|---|---|
| **Random variable** | A rule that assigns a numerical value to each outcome of a random process. |
| **Discrete random variable** | A random variable with a countable set of possible values (e.g., a count). |
| **Continuous random variable** | A random variable that can take any value within an interval (e.g., a measurement). |
| **Realization** | The specific observed value of a random variable after the process occurs. |
| **Probability distribution** | An assignment of probabilities to every possible value of a random variable; must be non-negative and sum to 1. |
| **Expected value** | $E(X) = \sum_x x\,P(X=x)$; the probability-weighted long-run average of $X$. |
| **Variance / standard deviation** | $\sigma^2 = \sum_x (x-\mu)^2 P(X=x)$; measures how spread out $X$'s values are around its mean. |

## Check Your Understanding

::: {.callout-note collapse="false" title="Problems"}
1. A ride-share company defines $X$ = number of cancellations per driver shift. Is $X$ discrete or continuous? What are some plausible values it could take?

2. A proposed distribution for the number of defective units in a shipment of 4 items is: $P(X=0)=0.6$, $P(X=1)=0.25$, $P(X=2)=0.1$, $P(X=3)=0.1$, $P(X=4)=0$. Is this a valid probability distribution? If not, explain why.

3. A subscription box company charges $30/month and estimates that 5% of customers will cancel and request a full refund each month, while the rest keep the box (no refund). Let $X$ = revenue per customer per month. Compute $E(X)$.

4. Using the distribution from Problem 3, would you expect every individual customer to generate close to $E(X)$ in a given month? Why or why not?
:::

::: {.callout-tip collapse="true" title="Solutions"}
1. $X$ is **discrete**&mdash;cancellations are counted, not measured. Plausible values are $0, 1, 2, 3, \dots$ up to however many rides a driver could plausibly be assigned in a shift.

2. Not valid. The probabilities sum to $0.6+0.25+0.1+0.1+0=1.05$, which exceeds 1, so this cannot be a real probability distribution (even though each individual value is between 0 and 1).

3. Let $X$ = revenue per customer. If the customer cancels: $X = \$0$ (full refund). If not: $X = \$30$.
   $$
   E(X) = (30)(0.95) + (0)(0.05) = 28.50
   $$
   Expected monthly revenue per customer is **$28.50**.

4. No. Every individual customer generates either exactly $0 or exactly $30&mdash;no one actually pays $28.50. The expected value is a long-run average across many customers, useful for forecasting total revenue across the subscriber base, not for predicting any one customer's outcome.
:::
