# Week 1: Importing Data, Identifying Variables, and Creating a Graph {#sec-week01-03}

## Why This Matters

Before any analytical technique is useful, you need to get a dataset into your software, understand what each row and column actually represents, and look at it visually. These three skills&mdash;**importing**, **identifying**, and **graphing**&mdash;are the mechanical starting point of every module in this course, so it's worth slowing down and doing them carefully here, before the pace picks up.

We'll work with a single running example throughout this reading: `readiness_store_sales.csv`, which tracks four months of performance across six retail stores.

| Column | Description |
|---|---|
| `store_id` | Unique identifier for the store |
| `region` | Region the store is located in (West, South, Midwest) |
| `month` | Month of the observation (Jan&ndash;Apr) |
| `sq_ft` | Store size, in square feet |
| `units_sold` | Number of units sold that month |
| `revenue` | Revenue that month, in dollars |

## Importing a Dataset

"Importing" simply means loading a data file&mdash;usually a CSV (comma-separated values) file&mdash;into the software you're going to analyze it with. The file itself doesn't change; you're just bringing a copy of it into Excel or R so you can work with it.

### In Excel

::: {.example title="Example 1.3: Opening a CSV in Excel"}
1. Locate `readiness_store_sales.csv` on your computer.

2. Open Excel, then use **File \> Open** and select the file (or simply double-click the file, which will open it directly in Excel).

3. Excel will display the file as a table: one column per variable, one row per observation, with the column names in the first row.

4. Select the data and choose **Home \> Format as Table** so Excel treats it as a proper table rather than a loose range of cells. This makes later steps&mdash;sorting, filtering, charting&mdash;much easier.
:::

### In R

In R, importing a dataset means reading the file into an object&mdash;essentially, giving the dataset a name so you can refer to it in later code. We'll use the `read_csv()` function from the **tidyverse**, a collection of R packages designed to work well together for data analysis.

::: {.example title="Example 1.4: Importing a CSV in R"}
```{r}
#| echo: true
#| message: false
library(tidyverse)

sales <- read_csv("data/readiness_store_sales.csv")
sales
```

The line `sales <- read_csv(...)` reads the file and stores it in an object named `sales`. Typing `sales` (or running a chunk containing just `sales`) prints it so you can see what was imported.
:::

::: {.callout-note collapse="false" title="A note on file paths"}
`read_csv("data/readiness_store_sales.csv")` tells R to look for a folder named `data` *inside your current project folder*, and for the file inside that folder. If R can't find the file, the most common cause is that your Quarto document isn't saved in the same project folder as the `data` folder. Keeping a consistent project folder structure now will save you real frustration later in the course.
:::

## Identifying the Observational Unit

Once a dataset is imported, the very first question to ask&mdash;before computing anything&mdash;is: **what does one row represent?** This is called the **observational unit** (sometimes called the *unit of analysis*). Every other decision about how to summarize, graph, or model the data depends on getting this right.

::: {.example title="Example 1.5: Finding the observational unit"}
In R, `glimpse()` is a quick way to see the shape of a dataset: how many rows and columns it has, along with each column's name, type, and a preview of its values.

```{r}
#| echo: true
glimpse(sales)
```

The output shows 24 rows. Looking at the data, each row corresponds to one combination of `store_id` and `month`&mdash;for example, store 101 in January is one row, and store 101 in February is a different row. So the observational unit here is **one store, in one month**&mdash;not "one store" (which would combine all four months together) and not "one month" (which would combine all six stores together).
:::

Getting the observational unit wrong is a common source of mistakes. If you mistakenly treated each row as "one store" (ignoring that stores repeat across months), you might average `revenue` down to one number per store-month combination and then, without realizing it, treat multiple rows for the same store as though they were independent stores.

## Identifying Variable Types

With the observational unit established, the next step is to classify each column. Two related distinctions matter most:

- **Categorical vs. quantitative:** Does the variable place observations into named groups (categorical), or is it measured on a numeric scale (quantitative)?
- **Discrete vs. continuous** (for quantitative variables only): Can the variable only take countable, distinct values (discrete), or can it take any value within a range (continuous)?

::: {.example title="Example 1.6: Classifying the columns of `sales`"}
| Column | Categorical or quantitative? | If quantitative: discrete or continuous? |
|---|---|---|
| `store_id` | Categorical (it labels a store; the numbers aren't meant to be added or averaged) | &mdash; |
| `region` | Categorical | &mdash; |
| `month` | Categorical | &mdash; |
| `sq_ft` | Quantitative | Continuous (in principle, a store could be any size, even though these happen to be round numbers) |
| `units_sold` | Quantitative | Discrete (you can't sell half a unit) |
| `revenue` | Quantitative | Continuous (dollars and cents can take any value within a range) |
:::

::: {.callout-note collapse="false" title="Watch out for numeric-looking categorical variables"}
`store_id` is stored as a number, and R's `glimpse()` may even show it with a numeric type. But storing something as a number doesn't make it quantitative. Ask yourself: *does it make sense to compute an average of this column?* The average of `revenue` is meaningful. The average of `store_id` is not&mdash;it's just a label. When in doubt, ask what the variable represents, not how it happens to be stored.
:::

## Creating One Graph

Once you know what a row represents and how each column is classified, you're ready to look at the data visually. A single, well-chosen graph is often more informative than a page of numbers.

### In Excel

::: {.example title="Example 1.7: A simple chart in Excel"}
1. Select the `month` and `revenue` columns (hold Ctrl, or Cmd on a Mac, to select two non-adjacent columns).

2. Go to **Insert \> Charts** and choose a line chart or clustered column chart.

3. Add a chart title and axis labels so the graph makes sense on its own, without needing the rest of the spreadsheet for context.
:::

### In R

We'll use `ggplot2` (part of the tidyverse) to build the graph. The basic idea behind `ggplot2` is that you map variables in your data to features of the graph&mdash;an x-axis, a y-axis, a color&mdash;and then add a "geometry" that says how to draw it (points, lines, bars, and so on).

::: {.example title="Example 1.8: A graph of revenue over time, by region"}
```{r}
#| echo: true
ggplot(sales, aes(x = month, y = revenue, color = region, group = store_id)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Monthly revenue by store",
    x = "Month",
    y = "Revenue ($)",
    color = "Region"
  )
```

A few things to notice about this code:

- `aes(x = month, y = revenue, color = region, group = store_id)` maps `month` to the horizontal axis, `revenue` to the vertical axis, `region` to color, and `store_id` to which points get connected by a line (since the observational unit is store-month, we need `group` to tell R to draw one line per store rather than connecting every point together).
- `geom_line()` and `geom_point()` add the visual elements&mdash;lines connecting each store's months, and points marking each observation.
- `labs()` adds a title and clear axis labels, which is what turns a plot into a *graph someone else can understand without you standing next to them*.
:::

::: {.callout-note collapse="false" title="Choosing what to graph"}
Because the observational unit is store-month, a graph of `revenue` by `month` naturally produces one line per store, as in Example 1.8. If you wanted a single overall trend instead, you would first need to summarize&mdash;for instance, adding up revenue across all six stores for each month&mdash;before graphing. We'll cover exactly this kind of summarizing in the next module.
:::

## Recap

- **Importing** a dataset brings a copy of a data file into your software (Excel or R) so you can work with it; it does not change the original file.
- The **observational unit** is what one row of the dataset represents. Identify it before doing anything else, since it shapes every summary, graph, and model that follows.
- Variables are **categorical** (naming groups) or **quantitative** (measured on a numeric scale); quantitative variables are further **discrete** (countable) or **continuous** (any value in a range). A variable's storage type (e.g., stored as a number) does not automatically make it quantitative&mdash;ask whether averaging it would be meaningful.
- A single, clearly labeled **graph** is often the fastest way to understand a dataset, but the right graph depends on knowing the observational unit and variable types first.

## Check Your Understanding

::: {.callout-note collapse="false" title="Problems"}
1. A hospital dataset has one row for every patient visit, with columns for `patient_id`, `department`, `visit_date`, `wait_time_minutes`, and `satisfaction_score` (1&ndash;5). What is the observational unit?

2. For the hospital dataset in Problem 1, classify each of the five columns as categorical or quantitative, and for the quantitative ones, as discrete or continuous.

3. Suppose you import a dataset in R and running `glimpse()` shows a column called `zip_code` stored as a numeric type. Should you treat `zip_code` as a quantitative variable? Why or why not?

4. You want to create a graph in R showing how `wait_time_minutes` relates to `department`. Which variable would you map to the x-axis, and which to the y-axis?
:::

::: {.callout-tip collapse="true" title="Solutions"}
1. The observational unit is **one patient visit**&mdash;not one patient (since a patient could have multiple visits) and not one department.

2. `patient_id`: categorical (a label, not a measurement). `department`: categorical. `visit_date`: categorical (it identifies when a visit occurred, similar to a label, rather than something you'd average). `wait_time_minutes`: quantitative, continuous (wait time could in principle be any value, e.g., 12.5 minutes). `satisfaction_score`: quantitative, discrete (only the whole values 1 through 5 are possible)&mdash;though in practice, some analysts treat scores like this as categorical (ordinal) as well, since the numeric distance between categories isn't always meaningful.

3. No. Even though `zip_code` is stored as a number, it functions as a label for a geographic area&mdash;computing an average zip code would be meaningless. It should be treated as categorical.

4. `department` (categorical) would typically go on the x-axis, and `wait_time_minutes` (quantitative) on the y-axis&mdash;for example, using a boxplot or bar chart to compare wait times across departments.
:::
