# Week 5: Confidence Intervals for a Mean {#sec-week05-02}

## From Proportions to Means

Last section built a confidence interval around $\hat{p}$ for a categorical outcome. This section does the same thing for a *quantitative* outcome, average order value, average handle time, average fill weight, using the sample mean $\bar{x}$. The overall shape is identical:
$$
\begin{align*}
\text{point estimate} \;&\pm\; \text{margin of error}\\
 \bar{x} \;&\pm\; \text{(critical value)} \times \text{standard error}
\end{align*}
$$
But there's a wrinkle here that didn't come up for proportions, and it's worth understanding *why* before we get to the formula.

## Estimating the Standard Error of $\bar{x}$

From @sec-week04-01, the standard error of $\bar{x}$ is $\sigma/\sqrt{n}$, where $\sigma$ is the population standard deviation. For proportions, the standard error formula, $\sqrt{p(1-p)/n}$, only involved the parameter $p$ we were already trying to estimate, so substituting $\hat{p}$ in its place was a clean fix. For a mean, $\sigma$ is a genuinely separate, unknown parameter, not something already tied to $\mu$. In practice, we estimate it with the sample standard deviation $s$, giving an estimated standard error:
$$
\text{SE}_{\bar{x}} = \frac{s}{\sqrt{n}}
$$

This substitution, using $s$ in place of the true $\sigma$, introduces extra uncertainty beyond what the normal distribution accounts for, especially when the sample is small. To handle that extra uncertainty honestly, we need a different reference distribution.

## The $t$-Distribution

The **$t$-distribution** looks like the standard normal distribution (bell-shaped, centered at 0), but with **heavier tails**, reflecting the added uncertainty of having estimated $\sigma$ with $s$ rather than knowing it exactly. How much heavier the tails are depends on the **degrees of freedom**, $df = n-1$: with a small sample, the tails are noticeably heavier than the normal curve, but as $n$ grows, the $t$-distribution converges to the standard normal.

```{r}
#| echo: false
#| message: false
library(tidyverse)

df_grid <- tibble(x = seq(-4, 4, length.out = 400)) %>%
  mutate(
    normal = dnorm(x),
    t_df5  = dt(x, df = 5),
    t_df20 = dt(x, df = 20)
  ) %>%
  pivot_longer(cols = c(normal, t_df5, t_df20), names_to = "dist", values_to = "density")

ggplot(df_grid, aes(x = x, y = density, color = dist)) +
  geom_line(linewidth = 1) +
  scale_color_manual(
    values = c(normal = "black", t_df5 = "#e6550d", t_df20 = "#3182bd"),
    labels = c("Standard normal", "t, df = 5", "t, df = 20")
  ) +
  labs(title = "The t-distribution converges to the normal as df grows",
       x = "Value", y = "Density", color = NULL)
```

Because its tails are heavier, using a $t$ critical value in place of a $z$ critical value produces a **wider** margin of error, appropriately wider, since we're accounting for the extra uncertainty of not knowing $\sigma$ exactly. With a small sample, this correction matters a lot; with a large sample, $t$ and $z$ critical values are nearly identical.

## Confidence Interval for a Mean

The confidence interval for $\mu$ uses a $t$ critical value, $t_{\alpha/2, \, df}$, with $df = n-1$:
$$
\bar{x} \;\pm\; t_{\alpha/2, \, df}\,\frac{s}{\sqrt{n}}
$$

To use this method, you need a randomly collected sample, and either an approximately normal population or a large enough $n$ for the Central Limit Theorem to make the sampling distribution of $\bar{x}$ approximately normal on its own.

::: {.callout-note collapse="false" title="This method is fairly robust to non-normality"}
The $t$-interval assumes the population is approximately normal, but in practice it performs reasonably well even when that assumption is modestly violated, especially as $n$ grows. The main exception is when the data contain extreme outliers: outliers distort both the sample mean itself and the standard deviation used to build the interval, so it's worth checking a histogram or boxplot for anything unusual before trusting the result, particularly with a small sample.
:::

::: {.example title="Example 5.3: A confidence interval for average order value"}
A retailer samples $n=25$ recent orders and finds a sample mean order value of $\bar{x}=\$71.50$ with a sample standard deviation of $s=\$42$. Build a 95% confidence interval for the true mean order value, $\mu$.

With $df = 25-1 = 24$, the 95% critical value is $t_{0.025, \, 24} = 2.064$ (found using software, shown below). The standard error is
$$
\text{SE}_{\bar{x}} = \frac{42}{\sqrt{25}} = 8.4
$$
The margin of error is
$$
2.064 \times 8.4 \approx 17.34
$$
The 95% confidence interval for $\mu$ is
$$
71.50 \pm 17.34, \quad \text{or} \quad (54.16, \, 88.84)
$$
We would report: "We are 95% confident that the true mean order value is between \$54.16 and \$88.84."
:::

## Computing $t$-Based Confidence Intervals in R and Excel

::: {.example title="Example 5.4: The order-value interval in R and Excel"}
```{r}
#| echo: true
x_bar <- 71.50
s <- 42
n <- 25
df <- n - 1

se <- s / sqrt(n)
t_star <- qt(0.975, df = df)

lower <- x_bar - t_star * se
upper <- x_bar + t_star * se

c(lower, upper)
```

`qt(0.975, df = 24)` plays the same role for the $t$-distribution that `qnorm(0.975)` played for the standard normal in @sec-week05-01. If you have the raw data rather than just the summary statistics, R's `t.test()` computes the same interval directly:
```{r}
#| echo: true
#| eval: false
t.test(order_values)$conf.int
```

In Excel, the critical value comes from `T.INV.2T()`, which returns the two-tailed critical value directly (no need to halve $\alpha$ yourself, unlike `NORM.S.INV()`):
```
=71.50 - T.INV.2T(0.05, 24) * (42/SQRT(25))   ' lower bound
=71.50 + T.INV.2T(0.05, 24) * (42/SQRT(25))   ' upper bound
```
:::

## When the Sample Is Large: $t$ vs. $z$

As degrees of freedom grow, the $t$-distribution's critical values get closer and closer to the corresponding $z$ critical values. By around $n=30$ or so, the difference is usually small enough not to matter much in practice, though using $t$ is always technically correct whenever $\sigma$ is estimated from the sample, which is nearly always the case in real business data.

## Recap

| **Keyword** | **Definition** |
|---|---|
| **Standard error of $\bar{x}$ (estimated)** | $\text{SE}_{\bar{x}} = s/\sqrt{n}$, using the sample standard deviation in place of the unknown $\sigma$. |
| **$t$-distribution** | A bell-shaped distribution with heavier tails than the normal, indexed by degrees of freedom; used when $\sigma$ is estimated by $s$. |
| **Degrees of freedom** | $df = n - 1$ for a confidence interval on a single mean. |
| **Confidence interval for $\mu$** | $\bar{x} \pm t_{\alpha/2, \, df}\,(s/\sqrt{n})$. |

## Check Your Understanding

::: {.callout-note collapse="false" title="Problems"}
1. A call center samples 16 calls and finds a mean handle time of $\bar{x}=7.2$ minutes with $s=2.1$ minutes. Using $t_{0.025, 15} = 2.131$, construct a 95% confidence interval for the true mean handle time.

2. Explain, in your own words, why the $t$-distribution has heavier tails than the standard normal distribution, and why that difference shrinks as $n$ grows.

3. A researcher mistakenly uses $z_{0.025}=1.96$ instead of the correct $t_{0.025, \, 9} = 2.262$ to build a confidence interval from a sample of $n=10$. Will the resulting interval be too wide or too narrow? Explain.

4. A production analyst samples 36 units and finds $\bar{x}=100.4$ with $s=5.0$. Using `qt(0.995, df = 35)` $\approx 2.724$, construct a 99% confidence interval for the true mean.
:::

::: {.callout-tip collapse="true" title="Solutions"}
1. $\text{SE} = 2.1/\sqrt{16} = 0.525$. Margin of error $= 2.131 \times 0.525 \approx 1.119$. The 95% confidence interval is $7.2 \pm 1.119$, or $(6.08, \, 8.32)$ minutes.

2. The $t$-distribution has heavier tails because it accounts for the extra uncertainty introduced by estimating $\sigma$ with $s$ rather than knowing it exactly, on top of the usual sampling variability in $\bar{x}$ itself. As $n$ grows, $s$ becomes a more precise estimate of $\sigma$, so that extra layer of uncertainty shrinks, and the $t$-distribution converges to the standard normal.

3. The interval will be **too narrow**. Because $z_{0.025}=1.96$ is smaller than the correct $t_{0.025,9}=2.262$, using it produces a smaller margin of error than the data actually justify, understating the true uncertainty and producing an interval less likely to actually capture $\mu$ 95% of the time.

4. $\text{SE} = 5.0/\sqrt{36} \approx 0.833$. Margin of error $= 2.724 \times 0.833 \approx 2.270$. The 99% confidence interval is $100.4 \pm 2.270$, or $(98.13, \, 102.67)$.
:::
